---
title: "simulating-survival-data"
output: html_document
date: "2025-04-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(rstatix)
library(survival)
library(survminer)
library(ggpubr)
```

```{python, setup}
import pandas as pd
import numpy as np
import lifelines
```

```{python, median plotting helper function}
def add_median_markers(fitters, ax=None, short_lines=True, colors=None):
    
    """
    Add median survival markers to a survival plot.

    Parameters:
    - fitters: A single fitter or a list of fitted lifelines KaplanMeierFitter models.
    - ax: Optional matplotlib Axes to draw into.
    - short_lines: If True (default), draw short lines. If False, draw full-width/height lines.
    - colors: Optional list of colors for the markers (must match number of fitters).
    """
    
    if not isinstance(fitters, list):
        fitters = [fitters]

    if ax is None:
        ax = plt.gca()

    if colors is None:
        # Cycle through default matplotlib colors
        colors = plt.rcParams['axes.prop_cycle'].by_key()['color']

    for i, kmf in enumerate(fitters):
        median = kmf.median_survival_time_
        color = colors[i % len(colors)]
        label = getattr(kmf, 'label', f'Group {i + 1}')

        if short_lines:
            # Short vertical and horizontal lines
            ax.plot([median, median], [0, 0.51], linestyle=':', color="gray", label=f'{label} Median: {median:.2f}')
            ax.plot([0, median], [0.5, 0.5], linestyle=':', color="gray")
        else:
            # Full lines across the plot
            ax.axvline(median, linestyle='--', color=color, label=f'{label} Median: {median:.2f}')
            ax.axhline(0.5, linestyle=':', color=color)

    ax.legend()
```


# Image settings

## Colours

For reference:

Orange - #EA7125 Blue - #0072CF Green - #55A51C Red - #D6083B Purple - #8F2BBC Cyan - #00B1C1

## ggplot settings

```{r}
theme_set(theme_classic(base_size = 15))

legend_base_size <- 14

output_dir <- "C:/Users/vicki/OneDrive/Documents/Bioinformatics/Survival-analysis/images/"
```

## ggsave settings

```{r}
base_height = 5
quarter_height = 2.5
base_width = 5
quarter_width = 2.5
ratio = 1.618

# required for survival curves specifically
ggsave_workaround <- function(g){survminer:::.build_ggsurvplot(x = g,
                                                               surv.plot.height = NULL,
                                                               risk.table.height = NULL,
                                                               ncensor.plot.height = NULL)}

# square plot
square.ggsave <- function(filename, height = base_height, width = base_width, ...) {
ggsave(filename = filename, height = height, width = width, ...)
}

# portrait plot
portrait.ggsave <- function(filename, height = base_height * ratio, width = base_width, ...) {
ggsave(filename = filename, height = height, width = width, ...)
}

# landscape plot
landscape.ggsave <- function(filename, height = base_height, width = base_width * ratio, ...) {
ggsave(filename = filename, height = height, width = width, ...)
}

# quarter plot
quarter.ggsave <- function(filename, height = quarter_height, width = quarter_width * ratio, ...) {
ggsave(filename = filename, height = height, width = width, ...)
}

# strip plot
strip.ggsave <- function(filename, height = base_height/3, width = base_width * 2, ...) {
ggsave(filename = filename, height = height, width = width, ...)
}

# wide plot
wide.ggsave <- function(filename, height = base_height, width = base_width * 2, ...) {
ggsave(filename = filename, height = height, width = width, ...)
}
```

# Simulate dataset

```{r}
set.seed(42)

# Sample size
n <- 300

# Define true coefficients (log-hazard ratios)
beta_age <- 0.03    # each year of age increases hazard
beta_sex <- 0.5     # being male increases hazard
```

The lin_pred is behaving like the y_avg that we use in the regular linear modelling. In a standard linear model, the link function is just the identity function, so the average predicted value of y is the same as the linear predictor. When we have a non-identity link function, the linear predictor will undergo the generalisation so that the mapping is no longer one-to-one.

We also need a baseline hazard function - this is sort of the baseline, a beta0.

```{r}
# Covariates
age <- rnorm(n, mean = 60, sd = 10)               # continuous
sex <- rbinom(n, 1, 0.5)                          # binary (0 = female, 1 = male)

# Linear predictor
lin_pred <- beta_age * age + beta_sex * sex

# Baseline hazard rate (lambda for exponential)
lambda <- 0.01
```

Now, we calculate the actual values of time using a transformation of that linear predictor.

Because the rexp function uses rate, we invert the hazard (use the exponential of lin_pred).

```{r}
# Simulate survival times
# T ~ Exponential(rate = lambda * exp(lin_pred))

time <- rexp(n, rate = lambda * exp(lin_pred))
```

Now, we simulate censoring, since survival data aren't just times.

First, create an object that simulates a bunch of cut-offs for censoring. These essentially reflect the last time that an individual could possibly have been observed in the study.

We use the uniform distribution, with a max value set to the 80th quantile of the `time` object we created just above.

```{r}
censoring <- runif(n, min = 0, max = quantile(time, 0.8))
```

The pmin function takes two vectors as input, and then returns a single vector containing the parllel minima. In other words, it takes each pair of values at each position (one from each vector) and returns whichever is the smallest.

Therefore, the `observed_time` object contains the last time the individual was actually observed. This will either be the final time they could possibly have been observed, taken from `censoring`, or the time that the event actually occurred.

Then, the `event` object contains just 0s and 1s, saying whether the event had occurred by that final censored time.

```{r}
observed_time <- pmin(time, censoring)

event <- as.numeric(time <= censoring)
```

Finally, we put all of this in a data frame.

These survival data are right-censored - the study cuts off before every individual has experienced the event of interest.

There is no left-censoring or interval-censoring here.

```{r}
observed_time <- round(observed_time)

surv_data <- data.frame(
  time = observed_time,
  event = event,
  age = age,
  sex = factor(sex, labels = c("Female", "Male"))
)

head(surv_data)
```

# Simulate a dragon hatching example using this process

```{r}
set.seed(15)

# Sample size
n <- 60

# Define true coefficients (log-hazard ratios)
beta_scales <- -0.5     # being metallic decreases hazard (increases hatching times)

# Covariates
scales <- rbinom(n, 1, 0.5)                          # binary (0 = chromatic, 1 = metallic)

# Linear predictor
lin_pred <- scales*beta_scales

# Baseline hazard rate (lambda for exponential)
lambda <- 0.03

# Simulate survival times
# T ~ Exponential(rate = lambda * exp(lin_pred))
time <- rexp(n, rate = lambda * exp(lin_pred))

censoring <- runif(n, min = 0, max = quantile(time, 0.8))
observed_time <- pmin(time, censoring)
event <- as.numeric(time <= censoring)

observed_time <- ceiling(observed_time) # using ceiling rather than round prevents left-censored events

hatching <- data.frame(
  time = observed_time,
  hatched = event,
  scales = factor(scales, labels = c("chromatic", "metallic"))
)

hatching <- hatching %>%
  mutate()

head(hatching)

write_csv(hatching, "C:/Users/vicki/Documents/BTF_repos/stats-survival-analysis/materials/data/hatching.csv")
```


# Survival curve for dragon hatching

## R

Survival curve for dragon egg hatching dataset (simulated above):

```{r}
hatch_curve <- survfit(Surv(time, event) ~ 1, hatching)
hatch_plot <- ggsurvplot(hatch_curve,
                         colour="black", palette="#EA7125",
                         ylab="Survival probability (egg intact)",
                         conf.int=TRUE,
                         font.legend = list(size = legend_base_size))

hatch_plot %>%
  ggsave_workaround() %>%
  landscape.ggsave(plot = ., filename = paste0(output_dir, "single_survival_curve.png"))
```

Curve with median line:

```{r}
hatch_curve <- survfit(Surv(time, event) ~ 1, hatching)
hatch_plot_median <- ggsurvplot(hatch_curve,
                         colour="black", palette="#EA7125",
                         ylab="Survival probability (egg intact)",
                         conf.int=TRUE,
                         surv.median.line="hv",
                         font.legend = list(size = legend_base_size))

hatch_plot_median %>%
  ggsave_workaround() %>%
  landscape.ggsave(plot = ., filename = paste0(output_dir, "survival_curve_withmedian.png"))
```

Curve with median line and risk table:

```{r}
hatch_curve <- survfit(Surv(time, event) ~ 1, hatching)
hatch_plot_table <- ggsurvplot(hatch_curve,
                         colour="black", palette="#EA7125",
                         ylab="Survival probability (egg intact)",
                         conf.int=TRUE,
                         surv.median.line="hv",
                         risk.table = TRUE,
                         font.legend = list(size = legend_base_size)) 

hatch_plot_table %>%
  ggsave_workaround() %>%
  landscape.ggsave(plot = ., filename = paste0(output_dir, "survival_curve_withtable.png"))
```

Curve with median line and cumulative events table:

```{r}
hatch_curve <- survfit(Surv(time, event) ~ 1, hatching)
hatch_plot_table2 <- ggsurvplot(hatch_curve,
                         colour="black", palette="#EA7125",
                         ylab="Survival probability (egg intact)",
                         conf.int=TRUE,
                         surv.median.line="hv",
                         cumevents = TRUE,
                         font.legend = list(size = legend_base_size)) 

hatch_plot_table2 %>%
  ggsave_workaround() %>%
  landscape.ggsave(plot = ., filename = paste0(output_dir, "survival_curve_witheventstable.png"))
```

## Python

```{python}
hatching = pd.read_csv("data/hatching.csv")
```

```{python}
hatch = KaplanMeierFitter()

T = hatching["time"]
E = hatching["hatched"]

hatch.fit(T, event_observed=E)
```

You could use `kmf.survival_function_.plot()`, but that's hijacking pandas and is much less customisable; better to use the `lifelines` function:

```{python}
plt.clf()
hatch.plot_survival_function()
plt.show()
```


Adding median line is a bit tricky. I've written a function, in the style of the `lifelines` package, that adds them for you.

```{python}
plt.clf()
hatch.plot_survival_function()
add_median_markers(hatch)
plt.show()
```

Adding risk tables can be done either via an optional argument in the plotting function, or (and this is better practice because it works for stratified curves), using the `add_counts` function.

```{python}
from lifelines.plotting import add_at_risk_counts

plt.clf()
hatch.plot_survival_function()
add_at_risk_counts(hatch)
add_median_markers(hatch)
plt.tight_layout()
plt.show()
```

# Stratified survival curve

## R

With risk table:

```{r}
scales_curve <- survfit(Surv(time, hatched) ~ scales, hatching)
hatch_plot_strat <- ggsurvplot(scales_curve, 
           palette=c("#0072CF","#EA7125"), 
           ylab="Survival probability (egg intact)", conf.int=TRUE,
           surv.median.line="hv", 
           risk.table=TRUE, tables.col = "strata",
           font.legend = list(size = legend_base_size))

hatch_plot_strat %>%
  ggsave_workaround() %>%
  landscape.ggsave(plot = ., filename = paste0(output_dir, "survival_curve_byscale.png"))
```

Without risk table:

```{r}
scales_curve <- survfit(Surv(time, event) ~ scales, hatching)
hatch_plot_strat_notab <- ggsurvplot(scales_curve, 
           palette=c("#0072CF","#EA7125"), 
           ylab="Survival probability (egg intact)", conf.int=TRUE,
           surv.median.line="hv", 
           tables.col = "strata",
           font.legend = list(size = text_base_size))

hatch_plot_strat_notab %>%
  ggsave_workaround() %>%
  landscape.ggsave(plot = ., filename = paste0(output_dir, "survival_curve_byscale_notable.png"))
```

## Python

```{python}
hatch = KaplanMeierFitter()

T = hatching["time"]
E = hatching["hatched"]
```

```{python}
plt.clf()

ax = plt.subplot(111)

met = (hatching["scales"] == "metallic")

met_hatch = KaplanMeierFitter()
met_hatch.fit(T[met], event_observed=E[met], label="metallic dragons")

chrom_hatch = KaplanMeierFitter()
chrom_hatch.fit(T[~met], event_observed=E[~met], label="chromatic dragons")

met_hatch.plot_survival_function(ax = ax)
chrom_hatch.plot_survival_function(ax = ax)

add_at_risk_counts(met_hatch, chrom_hatch, ax = ax)
add_median_markers([met_hatch, chrom_hatch], ax = ax)
plt.tight_layout()

plt.show()
```



# Mantel-Cox log hazards test

## R

```{r}
logrank_dragons <- survdiff(Surv(time, hatched) ~ scales, data=hatching)
```

```{r}
#| warning: false
scales_curve <- survfit(Surv(time, hatched) ~ scales, hatching)
hatch_plot_logrank <- ggsurvplot(scales_curve, 
           palette=c("#0072CF","#EA7125"), 
           ylab="Survival probability (egg intact)", conf.int=TRUE,
           surv.median.line="hv", 
           #risk.table=TRUE, tables.col = "strata",
           font.legend = list(size = legend_base_size),
           pval=TRUE, pval.method=TRUE)

hatch_plot_logrank %>%
  ggsave_workaround() %>%
  landscape.ggsave(plot = ., filename = paste0(output_dir, "survival_curve_stratwithp.png"))
```


## Python

```{python}
met = (hatching["scales"] == "metallic")

T_met, E_met = hatching.loc[met, 'time'], hatching.loc[met, 'hatched']
T_chrom, E_chrom = hatching.loc[~met, 'time'], hatching.loc[~met, 'hatched']

hatch_results = logrank_test(T_met, T_chrom, event_observed_A=E_met, event_observed_B=E_chrom)
hatch_results.print_summary()

print(hatch_results.test_statistic)
print(hatch_results.p_value)
```


# Simulating a dragon dataset with 3 groups

```{r}
#| warning: false
set.seed(15)

# Sample size
n <- 120

# Censoring quantile - the bigger this is, the less right-censoring there will be
cq <- 0.95

# End time of study (other source of right-censoring)
study_end <- 50

# Define true coefficients (log-hazard ratios)
beta_scales <- c(0, -1, 1) # blue as ref, gold is safer/slower, red is faster/less safe

# Covariates
scales <- sample(0:2, n, replace=TRUE) # 0=blue, 1=gold, 2=red

# Linear predictor
lin_pred <- scales*beta_scales

# Baseline hazard rate (lambda for exponential)
lambda <- 0.03

# Simulate survival times
# T ~ Exponential(rate = lambda * exp(lin_pred))
time <- rexp(n, rate = lambda * exp(lin_pred))

# create a random right-censor cut off for each individual
censoring <- runif(n, min = 0, max = quantile(time, cq))

# plus there is the study end to consider
total_censoring <- pmin(censoring, study_end)

# compare the censored cut offs and the event times for each individual
# take the smaller of the two to determine whether or not they were censored
observed_time <- pmin(time, total_censoring)

# and, if the censor time was the smaller of the two, they were censored
# so should be marked with a 0
# otherwise, mark with a 1
event <- as.numeric(time <= total_censoring)

observed_time <- ceiling(observed_time) # using ceiling rather than round prevents left-censored events

hatching3 <- data.frame(
  time = observed_time,
  hatched = event,
  scales = factor(scales, labels = c("blue", "gold", "red"))
)

scales_curve <- survfit(Surv(time, hatched) ~ scales, hatching3)
ggsurvplot(scales_curve, 
           palette=c("blue","gold", "red"), 
           ylab="Survival probability (egg intact)", conf.int=TRUE,
           surv.median.line="hv", 
           #risk.table=TRUE, tables.col = "strata",
           font.legend = list(size = legend_base_size),
           pval=TRUE, pval.method=TRUE)
```

```{r}
write_csv(hatching3, "C:/Users/vicki/Documents/BTF_repos/stats-survival-analysis/materials/data/hatching3.csv")
```


To reduce confidence intervals/p-values when making comparisons:

1. Increase sample size
2. Lowering censoring times: increase the event rate/reduce censoring
3. Increase baseline hazard (lambda)
4. Use more aggressive lin_pred/increase effect sizes between groups (beta)


# Log-rank test between three groups

```{r}
hatching3 <- read_csv("materials/data/hatching3.csv")
```

```{r}
survdiff(Surv(time, hatched) ~ scales, data=hatching3)
```

```{r}
#| warning: false
scales_curve <- survfit(Surv(time, hatched) ~ scales, hatching3)
hatch_plot_logrank3 <- ggsurvplot(scales_curve, 
           palette=c("blue","gold", "red"), 
           ylab="Survival probability (egg intact)", conf.int=TRUE,
           surv.median.line="hv", 
           #risk.table=TRUE, tables.col = "strata",
           font.legend = list(size = legend_base_size),
           pval=TRUE, pval.method=TRUE)

hatch_plot_logrank3
```

```{r}
hatch_plot_logrank %>%
  ggsave_workaround() %>%
  landscape.ggsave(plot = ., filename = paste0(output_dir, "survival_curve_stratwithp.png"))
```

Visualise this as three separate panels:

```{r}
#| warning: false
scales_curve <- survfit(Surv(time, hatched) ~ scales, hatching3)

hatch_plot_logrank_facet <- ggsurvplot(fit=scales_curve,
          palette=c("blue","gold", "red"), 
           ylab="Survival probability (egg intact)", conf.int=TRUE,
           surv.median.line="hv", 
           #risk.table=TRUE, tables.col = "strata",
           font.legend = list(size = legend_base_size))$plot +
  facet_wrap(~scales)

hatch_plot_logrank_facet
```


## Python

```{python}
hatching3 = pd.read_csv("materials/data/hatching3.csv")
```




# Fitting a Cox model

Fit a Cox model, using the `survival::coxph` function.

The syntax is really similar to regular linear modelling, except that the response variable is coded as `Surv(time, event)`.

This creates a "survival object" which we can look at more closely.

With this code, we're essentially telling R: "Model the hazard of the event as a function of age, sex, and treatment, using the survival time and event indicator from my data."

```{r}
Surv(time, event)

cox_fit <- coxph(Surv(time, event) ~ age + sex, data = surv_data)

summary(cox_fit)
```

If we want to extract p-values, we then need to do the proportional hazards test:

```{r}
# Proportional hazards test
cox.zph(cox_fit)
```

# Simulating interval censored data

If you want to have "realistic" survival data, they're probably going to be interval-censored
i.e., you know that the event occurred between the last time you looked and this time, but not exactly when.

To achieve this, you can either:

```{r}
# Round observed times to nearest integer
surv_data$time <- round(surv_data$time)  # Rounds to whole number (e.g., 1 unit = 1 month)

# Now there may be ties
table(duplicated(surv_data$time))  # TRUE = ties exist

```

Or;

```{r}
# Assume assessments happen monthly
visit_times <- seq(1, max(surv_data$time), by = 1)

# Round up to next visit time (ceiling)
surv_data$time <- ceiling(surv_data$time)

# Some extra ties now
head(table(surv_data$time))

```


rexp vs rgeom

rexp for continuous time
rgeom for stepwise time (e.g., observing once per month, or every morning)

## Seed germination example

```{r}
set.seed(123)

# 1. Parameters
n <- 300  # total seeds
species <- factor(sample(c("A", "B", "C"), size = n, replace = TRUE))  # 3 species

# 2. True species effects on germination (log-hazard scale)
beta <- c(A = 0, B = -0.5, C = 0.5)  # species B germinates slower, C faster

# 3. Convert to hazard rate via exp(beta)
hazard_rate <- exp(unname(beta[species])) * 0.2  # baseline daily hazard

# 4. Simulate true (unobserved) germination day
# T ~ Geometric(p = daily hazard) + 1 (to start at day 1)
true_day <- rgeom(n, prob = hazard_rate) + 1

# 5. Simulate censoring: e.g., study runs for 14 days
max_days <- 14
censored <- true_day > max_days
obs_day <- pmin(true_day, max_days)

# 6. Create interval-censored bounds
left <- ifelse(censored, max_days, obs_day - 1)
right <- ifelse(censored, NA, obs_day)

# 7. Final dataset
seed_data <- data.frame(
  species = species,
  left = left,
  right = right,
  germinated = !censored
)
```

```{r}
# Surv object with interval-censoring
S <- Surv(time = seed_data$left, time2 = seed_data$right, type = "interval2")

# Fit parametric survival model
fit <- survreg(S ~ species, data = seed_data, dist = "weibull")
summary(fit)
```

```{r}
# Convert right-censored observations to standard Surv object for plotting
plot_data <- seed_data
plot_data$time <- ifelse(plot_data$germinated, plot_data$right, plot_data$left)
plot_data$event <- as.numeric(plot_data$germinated)

ggsurvplot(survfit(Surv(time, event) ~ species, data = plot_data),
           conf.int = TRUE, pval = TRUE, risk.table = TRUE)
```


Cox model cannot handle interval censored data, which is why we used the parametric Weibull approach here.

That's because the Cox model, to estimate partial likelihood, requires us to know the exact order of when events occurred.

You can trick it, by using the midpoint of the interval as the exact time? But there's still going to be ties. ChatGPT didn't have a great explanation for why you might choose to trick it this way.

But there are other packages/functions (icenReg) that can do Cox regression on interval-censored data.

# Simulating hospital discharge dataset

```{r}
#| warning: false
# 20
set.seed(21)

n <- 200  # sample size

study_end <- 120

# Covariates
#age <- rnorm(n, mean = 45, sd = 15)
age <- rbeta(n, shape1 = 4, shape2 = 2) %>%
  scales::rescale(to = c(20, 85))
sex <- rbinom(n, 1, 0.5)
surgery <- rbinom(n, 1, 0.4)
weekend_admission <- rbinom(n, 1, 0.3)

# Log-hazard model
# Negative betas are protective against discharge (i.e., bad for the patient!)
beta <- c(
  age = -0.05,
  sex = 0.01,
  surgery = -0.5,
  weekend_admission = -0.3
)

lin_pred <- beta["age"] * age +
            beta["sex"] * sex +
            beta["surgery"] * surgery +
            beta["weekend_admission"] * weekend_admission

# Baseline hazard
lambda <- 1 # higher lambda → shorter average stay

# Simulate time-to-discharge
time_true <- rexp(n, rate = lambda * exp(lin_pred))

# Censoring
censoring <- runif(n, min = 0, max = study_end*3)
total_censoring <- pmin(censoring, study_end)
observed_time <- pmin(time_true, total_censoring)
event <- as.numeric(time_true <= total_censoring)

# Round to full days (optional)
observed_time <- round(observed_time) + 1

age <- round(age)
hospital_stays <- data.frame(
  time = observed_time/2,
  discharged = event,
  age, sex, surgery, weekend_admission
)


hosp_curve <- survfit(Surv(time, discharged) ~ 1, data = hospital_stays)
ggsurvplot(hosp_curve,
          colour = "black", palette = "#EA7125",
          ylab = "Probability of remaining in hospital",
          conf.int = TRUE,
          surv.median.line = "hv")

hosp_surg <- survfit(Surv(time, discharged) ~ surgery, data = hospital_stays)
ggsurvplot(hosp_surg,
          ylab = "Probability of remaining in hospital",
          conf.int = TRUE,
          surv.median.line = "hv",
          pval = TRUE)

hosp_wknd <- survfit(Surv(time, discharged) ~ weekend_admission, data = hospital_stays)
ggsurvplot(hosp_wknd,
          ylab = "Probability of remaining in hospital",
          conf.int = TRUE,
          surv.median.line = "hv",
          pval = TRUE)

hosp_sex <- survfit(Surv(time, discharged) ~ sex, data = hospital_stays)
ggsurvplot(hosp_sex,
          ylab = "Probability of remaining in hospital",
          conf.int = TRUE,
          surv.median.line = "hv",
          pval = TRUE)
```

```{r}
write_csv(hospital_stays, "materials/data/hospital_stays.csv")
```



```{r}
#| warning: false

pvals <- replicate(1000, {
  
  n <- 200  # sample size

  study_end <- 120
  cq <- 0.95
  
  # Covariates
  #age <- rnorm(n, mean = 45, sd = 15)
  age <- rbeta(n, shape1 = 4, shape2 = 2) %>%
    scales::rescale(to = c(20, 85))
  sex <- rbinom(n, 1, 0.5)
  surgery <- rbinom(n, 1, 0.44)
  weekend_admission <- rbinom(n, 1, 0.3)
  
  # Log-hazard model
  # Negative betas are protective against discharge (i.e., bad for the patient!)
  beta <- c(
    age = -0.05,
    sex = 0.01,
    surgery = -0.5,
    weekend_admission = -0.3
  )
  
  #lin_pred <- beta["age"] * age +
   # beta["sex"] * sex +
    #beta["surgery"] * surgery +
    #beta["weekend_admission"] * weekend_admission
  
  lin_pred <- beta["surgery"]*surgery
  
  # Baseline hazard
  lambda <- 0.1 # higher lambda → shorter average stay
  
  # Simulate time-to-discharge
  time_true <- rexp(n, rate = lambda * exp(lin_pred))
  
  # Censoring
  censoring <- runif(n, min = 0, max = study_end*3)
  total_censoring <- pmin(censoring, study_end)
  observed_time <- pmin(time_true, total_censoring)
  event <- as.numeric(time_true <= total_censoring)
  
  # Round to full days (optional)
  observed_time <- round(observed_time) + 1
  
  age <- round(age)
  hospital_stays <- data.frame(
    time = observed_time/2,
    discharged = event,
    age, sex, surgery, weekend_admission
  )
  
  surg_fit <- survdiff(Surv(time, discharged) ~ surgery, data = hospital_stays)
  surg_fit$pvalue
})

mean(pvals < 0.05)
```






