---
title: "simulating-survival-data"
output: html_document
date: "2025-04-16"
---

Ideas:

- Volcanic eruptions, earthquakes
- Eggs hatching (snakes?)
- Seeds germinating


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(rstatix)
library(survival)
library(survminer)
library(ggpubr)
```

```{python, setup}
import pandas as pd
import numpy as np
from lifelines import KaplanMeierFitter
kmf = KaplanMeierFitter()

```


# Image settings

## Colours

For reference:

Orange - #EA7125 Blue - #0072CF Green - #55A51C Red - #D6083B Purple - #8F2BBC Cyan - #00B1C1

## ggplot settings

```{r}
theme_set(theme_classic(base_size = 15))

legend_base_size <- 14

output_dir <- "C:/Users/vicki/OneDrive/Documents/Bioinformatics/Survival-analysis/images/"

setwd("C:/Users/vicki/Documents/BTF_repos/stats-survival-analysis")
```

## ggsave settings

```{r}
base_height = 5
quarter_height = 2.5
base_width = 5
quarter_width = 2.5
ratio = 1.618

# required for survival curves specifically
ggsave_workaround <- function(g){survminer:::.build_ggsurvplot(x = g,
                                                               surv.plot.height = NULL,
                                                               risk.table.height = NULL,
                                                               ncensor.plot.height = NULL)}

# square plot
square.ggsave <- function(filename, height = base_height, width = base_width, ...) {
ggsave(filename = filename, height = height, width = width, ...)
}

# portrait plot
portrait.ggsave <- function(filename, height = base_height * ratio, width = base_width, ...) {
ggsave(filename = filename, height = height, width = width, ...)
}

# landscape plot
landscape.ggsave <- function(filename, height = base_height, width = base_width * ratio, ...) {
ggsave(filename = filename, height = height, width = width, ...)
}

# quarter plot
quarter.ggsave <- function(filename, height = quarter_height, width = quarter_width * ratio, ...) {
ggsave(filename = filename, height = height, width = width, ...)
}

# strip plot
strip.ggsave <- function(filename, height = base_height/3, width = base_width * 2, ...) {
ggsave(filename = filename, height = height, width = width, ...)
}

# wide plot
wide.ggsave <- function(filename, height = base_height, width = base_width * 2, ...) {
ggsave(filename = filename, height = height, width = width, ...)
}
```

# Simulate dataset

```{r}
set.seed(42)

# Sample size
n <- 300

# Define true coefficients (log-hazard ratios)
beta_age <- 0.03    # each year of age increases hazard
beta_sex <- 0.5     # being male increases hazard
```

The lin_pred is behaving like the y_avg that we use in the regular linear modelling. In a standard linear model, the link function is just the identity function, so the average predicted value of y is the same as the linear predictor. When we have a non-identity link function, the linear predictor will undergo the generalisation so that the mapping is no longer one-to-one.

We also need a baseline hazard function - this is sort of the baseline, a beta0.

```{r}
# Covariates
age <- rnorm(n, mean = 60, sd = 10)               # continuous
sex <- rbinom(n, 1, 0.5)                          # binary (0 = female, 1 = male)

# Linear predictor
lin_pred <- beta_age * age + beta_sex * sex

# Baseline hazard rate (lambda for exponential)
lambda <- 0.01
```

Now, we calculate the actual values of time using a transformation of that linear predictor.

Because the rexp function uses rate, we invert the hazard (use the exponential of lin_pred).

```{r}
# Simulate survival times
# T ~ Exponential(rate = lambda * exp(lin_pred))

time <- rexp(n, rate = lambda * exp(lin_pred))
```

Now, we simulate censoring, since survival data aren't just times.

First, create an object that simulates a bunch of cut-offs for censoring. These essentially reflect the last time that an individual could possibly have been observed in the study.

We use the uniform distribution, with a max value set to the 80th quantile of the `time` object we created just above.

```{r}
censoring <- runif(n, min = 0, max = quantile(time, 0.8))
```

The pmin function takes two vectors as input, and then returns a single vector containing the parllel minima. In other words, it takes each pair of values at each position (one from each vector) and returns whichever is the smallest.

Therefore, the `observed_time` object contains the last time the individual was actually observed. This will either be the final time they could possibly have been observed, taken from `censoring`, or the time that the event actually occurred.

Then, the `event` object contains just 0s and 1s, saying whether the event had occurred by that final censored time.

```{r}
observed_time <- pmin(time, censoring)

event <- as.numeric(time <= censoring)
```

Finally, we put all of this in a data frame.

These survival data are right-censored - the study cuts off before every individual has experienced the event of interest.

There is no left-censoring or interval-censoring here.

```{r}
observed_time <- round(observed_time)

surv_data <- data.frame(
  time = observed_time,
  event = event,
  age = age,
  sex = factor(sex, labels = c("Female", "Male"))
)

head(surv_data)
```

# Simulate a dragon hatching example using this process

```{r}
set.seed(15)

# Sample size
n <- 60

# Define true coefficients (log-hazard ratios)
beta_scales <- -0.5     # being metallic decreases hazard (increases hatching times)

# Covariates
scales <- rbinom(n, 1, 0.5)                          # binary (0 = chromatic, 1 = metallic)

# Linear predictor
lin_pred <- scales*beta_scales

# Baseline hazard rate (lambda for exponential)
lambda <- 0.03

# Simulate survival times
# T ~ Exponential(rate = lambda * exp(lin_pred))
time <- rexp(n, rate = lambda * exp(lin_pred))

censoring <- runif(n, min = 0, max = quantile(time, 0.8))
observed_time <- pmin(time, censoring)
event <- as.numeric(time <= censoring)

observed_time <- ceiling(observed_time) # using ceiling rather than round prevents left-censored events

hatching <- data.frame(
  time = observed_time,
  hatched = event,
  scales = factor(scales, labels = c("chromatic", "metallic"))
)

hatching <- hatching %>%
  mutate()

head(hatching)

write_csv(hatching, "C:/Users/vicki/Documents/BTF_repos/stats-survival-analysis/materials/data/hatching.csv")
```

# Survival curve: population vs estimate

```{r}
set.seed(115)

# Sample size
n <- 60

# No covariates → single population
lin_pred <- 0

# Baseline hazard
lambda <- 0.05

# Survival times
time <- rexp(n, rate = lambda * exp(lin_pred))

# Censoring
censoring <- runif(n, min = 0, max = quantile(time, 0.8))
observed_time <- pmin(time, censoring)
event <- as.numeric(time <= censoring)

observed_time <- ceiling(observed_time)

data <- data.frame(
  time = observed_time,
  event = event
)
```

```{r}
# Time grid
t_grid <- seq(0, max(data$time), length.out = 200)

# True survival
S_true <- exp(-lambda * t_grid)

true_df <- data.frame(
  time = t_grid,
  survival = S_true
)

# Plot
true_plot <- ggplot(true_df, aes(x = time, y = survival)) +
  geom_line(linewidth = 1.2, color = "#00B1C1") +
  labs(title = "True survival curve",
       x = "Time",
       y = "S(t)") +
  theme_classic(base_size = 15)
```

```{r}
# KM fit
fit <- survfit(Surv(time, event) ~ 1, data = data)

# Plot
km_plot <- ggsurvplot(
  fit,
  data = data,
  conf.int = TRUE,
  title = "Estimated survival curve",
  palette = "#EA7125",
  colour = "#00B1C1",
)
```

```{r}
x_max <- max(data$time)

true_plot <- true_plot +
  coord_cartesian(xlim = c(0, x_max), ylim = c(0, 1))

km_plot$plot <- km_plot$plot +
  theme_classic(base_size = 15) +
  coord_cartesian(xlim = c(0, x_max), ylim = c(0, 1)) + theme(legend.position = "none")

true_est_surv_curve <- true_plot | km_plot$plot
true_est_surv_curve
```

```{r}
ggsave("materials/images/km_est_plot.png", km_plot$plot, width = 4.75, height = 5)
ggsave("materials/images/true_plot.png", true_plot, width = 4.75, height = 5)
ggsave("materials/images/true_est_plot.png", true_est_surv_curve, width = 9.5, height = 5)
```


# Survival curve for dragon hatching

Read in data - is simulated above.

```{r}
hatching <- read_csv("C:/Users/vicki/Documents/BTF_repos/stats-survival-analysis/materials/data/hatching.csv")
```

## Previous attempts

Survival curve for dragon egg hatching dataset (simulated above):
```{r}
hatch_curve <- survfit(Surv(time, hatched) ~ 1, hatching)

hatch_plot <- ggsurvplot(hatch_curve)

hatch_plot <- ggsurvplot(hatch_curve,
                         colour="black", palette="#EA7125",
                         ylab="Survival probability (egg intact)",
                         conf.int=TRUE,
                         font.legend = list(size = legend_base_size))

hatch_plot %>%
  ggsave_workaround() %>%
  landscape.ggsave(plot = ., filename = paste0(output_dir, "single_survival_curve.png"))
```

Curve with median line:
```{r}
hatch_curve <- survfit(Surv(time, event) ~ 1, hatching)

hatch_plot_median <- ggsurvplot(hatch_curve)

hatch_plot_median <- ggsurvplot(hatch_curve,
                         colour="black", palette="#EA7125",
                         ylab="Survival probability (egg intact)",
                         conf.int=TRUE,
                         surv.median.line="hv",
                         font.legend = list(size = legend_base_size))

hatch_plot_median %>%
  ggsave_workaround() %>%
  landscape.ggsave(plot = ., filename = paste0(output_dir, "survival_curve_withmedian.png"))
```

Curve with median line and risk table:
```{r}
hatch_curve <- survfit(Surv(time, event) ~ 1, hatching)
hatch_plot_table <- ggsurvplot(hatch_curve,
                         colour="black", palette="#EA7125",
                         ylab="Survival probability (egg intact)",
                         conf.int=TRUE,
                         surv.median.line="hv",
                         risk.table = TRUE,
                         font.legend = list(size = legend_base_size)) 

hatch_plot_table %>%
  ggsave_workaround() %>%
  landscape.ggsave(plot = ., filename = paste0(output_dir, "survival_curve_withtable.png"))
```

Curve with median line and cumulative events table:
```{r}
hatch_curve <- survfit(Surv(time, event) ~ 1, hatching)
hatch_plot_table2 <- ggsurvplot(hatch_curve,
                         colour="black", palette="#EA7125",
                         ylab="Survival probability (egg intact)",
                         conf.int=TRUE,
                         surv.median.line="hv",
                         cumevents = TRUE,
                         font.legend = list(size = legend_base_size)) 

hatch_plot_table2 %>%
  ggsave_workaround() %>%
  landscape.ggsave(plot = ., filename = paste0(output_dir, "survival_curve_witheventstable.png"))
```

## Stratified survival curve

With risk table:
```{r}
scales_curve <- survfit(Surv(time, hatched) ~ scales, hatching)
hatch_plot_strat <- ggsurvplot(scales_curve, 
           palette=c("#0072CF","#EA7125"), 
           ylab="Survival probability (egg intact)", conf.int=TRUE,
           surv.median.line="hv", 
           risk.table=TRUE, tables.col = "strata",
           font.legend = list(size = text_base_size))

hatch_plot_strat %>%
  ggsave_workaround() %>%
  landscape.ggsave(plot = ., filename = paste0(output_dir, "survival_curve_byscale.png"))
```

Without risk table:
```{r}
scales_curve <- survfit(Surv(time, event) ~ scales, hatching)
hatch_plot_strat_notab <- ggsurvplot(scales_curve, 
           palette=c("#0072CF","#EA7125"), 
           ylab="Survival probability (egg intact)", conf.int=TRUE,
           surv.median.line="hv", 
           tables.col = "strata",
           font.legend = list(size = text_base_size))

hatch_plot_strat_notab %>%
  ggsave_workaround() %>%
  landscape.ggsave(plot = ., filename = paste0(output_dir, "survival_curve_byscale_notable.png"))
```

## Running log rank test

```{r}
scales_curve <- survfit(Surv(time, event) ~ scales, hatching)

dragon_curve1 <- ggsurvplot(scales_curve, conf.int = TRUE,
            palette=c("#8F2BBC","#55A51C"), conf.int.alpha = 0.2,
            ylab="Survival probability (egg intact)",
            title = "Dragon egg hatching", subtitle = "Comparing two curves",
            legend = "right", legend.title = "Scale colour", legend.labs = c("Chromatic", "Metallic"))

dragon_curve2 <- ggsurvplot(scales_curve, conf.int = TRUE, pval = TRUE,
            surv.median.line = "hv",
            palette=c("#8F2BBC","#55A51C"), conf.int.alpha = 0.2,
            ylab="Survival probability (egg intact)",
            title = "Dragon egg hatching", subtitle = "Comparing two curves",
            legend = "right", legend.title = "Scale colour", legend.labs = c("Chromatic", "Metallic"))
```

```{r}
dragon_curve1 %>%
  ggsave_workaround() %>%
  landscape.ggsave(plot = ., filename = "C:/Users/vicki/Documents/BTF_repos/stats-survival-analysis/materials/images/dragon_curve1.png")

dragon_curve2 %>%
  ggsave_workaround() %>%
  landscape.ggsave(plot = ., filename = "C:/Users/vicki/Documents/BTF_repos/stats-survival-analysis/materials/images/dragon_curve2.png")
```

## Three curves

```{r}
# Add a third group of dragons to the hatching dataset via simulation
set.seed(15)

# Sample size
n <- 90

# Define true coefficients (log-hazard ratios)
beta_scales <- c(-0.5, 0.5)     # being metallic decreases hazard (increases hatching times), being iridescent increases hazard (decreases hatching times)

# Create scales variable
scales <- sample(c("chromatic", "metallic", "iridescent"), size = n, replace = TRUE)

# Linear predictor
lin_pred <- ifelse(scales == "chromatic", 0, ifelse(scales == "metallic", beta_scales[1], beta_scales[2]))

# Baseline hazard
lambda <- 0.03

# Simulate survival times
# T ~ Exponential(rate = lambda * exp(lin_pred))
time <- rexp(n, rate = lambda * exp(lin_pred))

censoring <- runif(n, min = 0, max = quantile(time, 0.8))
observed_time <- pmin(time, censoring)
event <- as.numeric(time <= censoring)

observed_time <- ceiling(observed_time) # using ceiling rather than round prevents left-censored events

hatching3 <- data.frame(
  time = observed_time,
  hatched = event,
  scales = factor(scales, labels = c("chromatic", "metallic", "iridescent"))
)

# Take the iridescent dragons from hatching3 and join them with the original hatching dataset
hatching3_exact <- hatching3 %>%
  filter(scales == "iridescent") %>%
  select(time, hatched, scales) %>%
  bind_rows(hatching) %>%
  # convert scales to factor in c, m, i order
  mutate(scales = factor(scales, levels = c("chromatic", "metallic", "iridescent")))
  
```

```{r}
scales_curve_3 <- survfit(Surv(time, hatched) ~ scales, hatching3_exact)

dragon_curve3 <- ggsurvplot(scales_curve_3, conf.int = TRUE,
            palette=c("#8F2BBC","#55A51C", "#EA7125"), conf.int.alpha = 0.2,
            ylab="Survival probability (egg intact)", xlim = c(0, 50),
            title = "Dragon egg hatching", subtitle = "Comparing three curves",
            legend = "right", legend.title = "Scale colour", legend.labs = c("Chromatic", "Metallic", "Iridescent"))

dragon_curve4 <- ggsurvplot(scales_curve_3, conf.int = TRUE, pval = TRUE,
            palette=c("#8F2BBC", "#55A51C", "#EA7125"), conf.int.alpha = 0.2,
            surv.median.line = "hv",
            ylab="Survival probability (egg intact)", xlim = c(0, 50),
            title = "Dragon egg hatching", subtitle = "Comparing three curves",
            legend = "right", legend.title = "Scale colour", legend.labs = c("Chromatic", "Metallic", "Iridescent"))
```

```{r}
dragon_curve3 %>%
  ggsave_workaround() %>%
  landscape.ggsave(plot = ., filename = "C:/Users/vicki/Documents/BTF_repos/stats-survival-analysis/materials/images/dragon_curve3.png")

dragon_curve4 %>%
  ggsave_workaround() %>%
  landscape.ggsave(plot = ., filename = "C:/Users/vicki/Documents/BTF_repos/stats-survival-analysis/materials/images/dragon_curve4.png")
```

# Mantel-Cox log hazards test

```{r}
survdiff(Surv(time, event) ~ sex, data=surv_data)
```


# Visualising theoretical survival curves in log rank test

Null is true:

```{r}
set.seed(71)

# Sample size
n <- 60

# No effect
beta_scales <- 0

# Covariates
scales <- rbinom(n, 1, 0.5)

# Linear predictor
lin_pred <- scales * beta_scales

# Baseline hazard
lambda <- 0.03

# Survival times
time <- rexp(n, rate = lambda * exp(lin_pred))

# Censoring
censoring <- runif(n, min = 0, max = quantile(time, 0.8))
observed_time <- pmin(time, censoring)
event <- as.numeric(time <= censoring)

observed_time <- ceiling(observed_time)

hatching_null <- data.frame(
  time = observed_time,
  hatched = event,
  scales = factor(scales, labels = c("chromatic", "metallic"))
)

# KM fit
fit_null <- survfit(Surv(time, hatched) ~ scales, data = hatching_null)

# Log-rank test
lr_null <- survdiff(Surv(time, hatched) ~ scales, data = hatching_null)
p_null <- 1 - pchisq(lr_null$chisq, df = 1)

# Plot
h0_survplot <- ggsurvplot(
  fit_null,
  data = hatching_null,
  pval = FALSE,
  legend="none",
  title = "Null hypothesis is true"
)

h0_survplot <- h0_survplot$plot + 
  coord_cartesian(xlim = c(0,50)) + 
  theme_classic(base_size = 15) + 
  theme(legend.position = "none")
```

Null is false: 

```{r}
set.seed(71)

# Sample size
n <- 60

# Effect present
beta_scales <- -0.8   # stronger effect for clearer separation

# Covariates
scales <- rbinom(n, 1, 0.5)

# Linear predictor
lin_pred <- scales * beta_scales

# Baseline hazard
lambda <- 0.03

# Survival times
time <- rexp(n, rate = lambda * exp(lin_pred))

# Censoring
censoring <- runif(n, min = 0, max = quantile(time, 0.8))
observed_time <- pmin(time, censoring)
event <- as.numeric(time <= censoring)

observed_time <- ceiling(observed_time)

hatching_diff <- data.frame(
  time = observed_time,
  hatched = event,
  scales = factor(scales, labels = c("chromatic", "metallic"))
)

# KM fit
fit_diff <- survfit(Surv(time, hatched) ~ scales, data = hatching_diff)

# Log-rank test
lr_diff <- survdiff(Surv(time, hatched) ~ scales, data = hatching_diff)
p_diff <- 1 - pchisq(lr_diff$chisq, df = 1)

# Plot
h1_survplot <- ggsurvplot(
  fit_diff,
  data = hatching_diff,
  pval = FALSE,
  legend="none",
  title = "Null hypothesis is false"
) 

h1_survplot <- h1_survplot$plot + 
  coord_cartesian(xlim = c(0,80)) + 
  theme_classic(base_size = 15) + 
  theme(legend.position = "none")
```

```{r}
log_rank_h0_h1 <- h0_survplot + h1_survplot

log_rank_h0_h1
```

```{r}
ggsave("materials/images/log_rank_h0_h1.png", plot = log_rank_h0_h1,
       width = 9, height = 4)
```

## Produce the "real" survival curves

```{r}
library(ggplot2)

# Time grid
t_grid <- seq(0, max(hatching_diff$time), length.out = 200)

# True survival
S_chromatic <- exp(-lambda * t_grid)
S_metallic  <- exp(-lambda * exp(beta_scales) * t_grid)

true_diff <- data.frame(
  time = rep(t_grid, 2),
  survival = c(S_chromatic, S_metallic),
  scales = factor(rep(c("chromatic", "metallic"), each = length(t_grid)))
)

# Plot
surv_curves_h1 <- ggplot(true_diff, aes(x = time, y = survival, color = scales)) +
  geom_line(linewidth = 1.2) +
  labs(title = "Null hypothesis is false",
       x = "Time",
       y = "S(t)") +
  theme_classic() +
  coord_cartesian(ylim = c(0,1), xlim = c(0,80)) +
  theme_classic(base_size = 15) + theme(legend.position = "none")

```

```{r}
# Time grid
t_grid <- seq(0, max(hatching_null$time), length.out = 200)

epsilon <- 0.0075  # very small shift

S_metallic_offset <- pmin(exp(-lambda * t_grid) + epsilon, 1)

true_null_offset <- data.frame(
  time = rep(t_grid, 2),
  survival = c(exp(-lambda * t_grid), S_metallic_offset),
  scales = factor(rep(c("chromatic", "metallic"), each = length(t_grid)))
)

surv_curves_h0 <- ggplot(true_null_offset, aes(x = time, y = survival, color = scales)) +
  geom_line(linewidth = 1.2) +
  labs(title = "Null hypothesis is true",
       x = "Time",
       y = "S(t)") +
  theme_classic() +
  coord_cartesian(ylim = c(0,1), xlim = c(0,50)) +
  theme_classic(base_size = 15) + theme(legend.position = "none")
```

```{r}
surv_curves_h0_h1 <- surv_curves_h0 + surv_curves_h1 
surv_curves_h0_h1
```

```{r}
ggsave("materials/images/surv_curves_h0_h1.png", plot = surv_curves_h0_h1,
       width = 9, height = 4)
```

# Fitting a Cox model

Fit a Cox model, using the `survival::coxph` function.

The syntax is really similar to regular linear modelling, except that the response variable is coded as `Surv(time, event)`.

This creates a "survival object" which we can look at more closely.

With this code, we're essentially telling R: "Model the hazard of the event as a function of age, sex, and treatment, using the survival time and event indicator from my data."

```{r}
Surv(time, event)

cox_fit <- coxph(Surv(time, event) ~ age + sex, data = surv_data)

summary(cox_fit)
```

If we want to extract p-values, we then need to do the proportional hazards test:

```{r}
# Proportional hazards test
cox.zph(cox_fit)
```

# Simulating interval censored data

If you want to have "realistic" survival data, they're probably going to be interval-censored
i.e., you know that the event occurred between the last time you looked and this time, but not exactly when.

To achieve this, you can either:

```{r}
# Round observed times to nearest integer
surv_data$time <- round(surv_data$time)  # Rounds to whole number (e.g., 1 unit = 1 month)

# Now there may be ties
table(duplicated(surv_data$time))  # TRUE = ties exist

```

Or;

```{r}
# Assume assessments happen monthly
visit_times <- seq(1, max(surv_data$time), by = 1)

# Round up to next visit time (ceiling)
surv_data$time <- ceiling(surv_data$time)

# Some extra ties now
head(table(surv_data$time))

```


rexp vs rgeom

rexp for continuous time
rgeom for stepwise time (e.g., observing once per month, or every morning)

## Seed germination example

```{r}
set.seed(123)

# 1. Parameters
n <- 300  # total seeds
species <- factor(sample(c("A", "B", "C"), size = n, replace = TRUE))  # 3 species

# 2. True species effects on germination (log-hazard scale)
beta <- c(A = 0, B = -0.5, C = 0.5)  # species B germinates slower, C faster

# 3. Convert to hazard rate via exp(beta)
hazard_rate <- exp(unname(beta[species])) * 0.2  # baseline daily hazard

# 4. Simulate true (unobserved) germination day
# T ~ Geometric(p = daily hazard) + 1 (to start at day 1)
true_day <- rgeom(n, prob = hazard_rate) + 1

# 5. Simulate censoring: e.g., study runs for 14 days
max_days <- 14
censored <- true_day > max_days
obs_day <- pmin(true_day, max_days)

# 6. Create interval-censored bounds
left <- ifelse(censored, max_days, obs_day - 1)
right <- ifelse(censored, NA, obs_day)

# 7. Final dataset
seed_data <- data.frame(
  species = species,
  left = left,
  right = right,
  germinated = !censored
)
```

```{r}
# Surv object with interval-censoring
S <- Surv(time = seed_data$left, time2 = seed_data$right, type = "interval2")

# Fit parametric survival model
fit <- survreg(S ~ species, data = seed_data, dist = "weibull")
summary(fit)
```

```{r}
# Convert right-censored observations to standard Surv object for plotting
plot_data <- seed_data
plot_data$time <- ifelse(plot_data$germinated, plot_data$right, plot_data$left)
plot_data$event <- as.numeric(plot_data$germinated)

ggsurvplot(survfit(Surv(time, event) ~ species, data = plot_data),
           conf.int = TRUE, pval = TRUE, risk.table = TRUE)
```


Cox model cannot handle interval censored data, which is why we used the parametric Weibull approach here.

That's because the Cox model, to estimate partial likelihood, requires us to know the exact order of when events occurred.

You can trick it, by using the midpoint of the interval as the exact time? But there's still going to be ties. ChatGPT didn't have a great explanation for why you might choose to trick it this way.

But there are other packages/functions (icenReg) that can do Cox regression on interval-censored data.


# Visualising (non-)proportional hazards

```{r}
set.seed(71)

library(survival)

n <- 500
```

```{r}
sim_ph <- function(n, lambda = 0.10, HR = 0.60, censor_rate = 0.02) {

  trt <- rbinom(n, 1, 0.5)

  # Constant hazard ratio:
  # treatment hazard = lambda * HR
  hazard <- lambda * ifelse(trt == 1, HR, 1)

  event_time <- rexp(n, rate = hazard)
  censor_time <- rexp(n, rate = censor_rate)

  time <- pmin(event_time, censor_time)
  status <- as.integer(event_time <= censor_time)

  data.frame(time, status, trt)
}
```

Crossing hazards
```{r}
sim_crossing <- function(n,
                         lambda_control = 0.10,
                         lambda_trt_early = 0.18,
                         lambda_trt_late = 0.04,
                         change_time = 5,
                         censor_rate = 0.01) {

  trt <- rbinom(n, 1, 0.5)

  event_time <- numeric(n)

  for (i in seq_len(n)) {

    if (trt[i] == 0) {

      # Control: constant hazard
      event_time[i] <- rexp(1, rate = lambda_control)

    } else {

      # Treatment: high hazard initially
      t1 <- rexp(1, rate = lambda_trt_early)

      if (t1 <= change_time) {
        event_time[i] <- t1
      } else {
        # Survived the early period; generate remaining time
        event_time[i] <-
          change_time +
          rexp(1, rate = lambda_trt_late)
      }
    }
  }

  censor_time <- rexp(n, rate = censor_rate)

  time <- pmin(event_time, censor_time)
  status <- as.integer(event_time <= censor_time)

  data.frame(time, status, trt)
}
```

Delayed effect
```{r}
sim_delayed <- function(n,
                        lambda = 0.10,
                        HR_after = 0.40,
                        delay = 5,
                        censor_rate = 0.01) {

  trt <- rbinom(n, 1, 0.5)

  event_time <- numeric(n)

  for (i in seq_len(n)) {

    if (trt[i] == 0) {

      event_time[i] <- rexp(1, rate = lambda)

    } else {

      # No treatment effect before the delay
      t1 <- rexp(1, rate = lambda)

      if (t1 <= delay) {
        event_time[i] <- t1
      } else {
        event_time[i] <-
          delay +
          rexp(1, rate = lambda * HR_after)
      }
    }
  }

  censor_time <- rexp(n, rate = censor_rate)

  time <- pmin(event_time, censor_time)
  status <- as.integer(event_time <= censor_time)

  data.frame(time, status, trt)
}
```

Cure effect function
```{r}
sim_cure <- function(n,
                     lambda = 0.10,
                     cure_treatment = 0.35,
                     censor_rate = 0.01) {

  trt <- rbinom(n, 1, 0.5)

  # Only the treatment group can be cured
  cure_prob <- ifelse(
    trt == 1,
    cure_treatment,
    0
  )

  cured <- rbinom(n, 1, cure_prob)

  # Event times for susceptible subjects
  event_time <- rexp(n, rate = lambda)

  # Cured treatment subjects never experience the event
  event_time[cured == 1] <- Inf

  # Independent censoring
  censor_time <- rexp(n, rate = censor_rate)

  time <- pmin(event_time, censor_time)
  status <- as.integer(event_time <= censor_time)

  data.frame(
    time,
    status,
    trt,
    cured
  )
}
```

Increasing hazards
```{r}
sim_increasing_effect <- function(
    n,
    lambda = 0.10,
    HR_start = 1.0,
    HR_end = 0.30,
    scale = 10,
    censor_rate = 0.01
) {

  trt <- rbinom(n, 1, 0.5)

  # Smoothly changing HR:
  # HR_start at t = 0
  # approaches HR_end as t increases
  HR <- function(t) {
    HR_end + (HR_start - HR_end) * exp(-t / scale)
  }

  # Cumulative treatment hazard
  H_trt <- function(t) {
    lambda * (
      HR_end * t +
      (HR_start - HR_end) * scale *
        (1 - exp(-t / scale))
    )
  }

  event_time <- numeric(n)

  for (i in seq_len(n)) {

    u <- runif(1)
    target <- -log(u)

    if (trt[i] == 0) {

      # Control: exponential survival
      event_time[i] <- target / lambda

    } else {

      # Find t such that H(t) = -log(U)
      event_time[i] <- uniroot(
        function(t) H_trt(t) - target,
        interval = c(0, 1e4)
      )$root
    }
  }

  censor_time <- rexp(n, rate = censor_rate)

  time <- pmin(event_time, censor_time)
  status <- as.integer(event_time <= censor_time)

  data.frame(time, status, trt)
}
```

Decreasing hazards
```{r}
sim_decreasing_effect <- function(
    n,
    lambda = 0.10,
    HR_start = 0.30,
    HR_end = 1.0,
    scale = 8,
    censor_rate = 0.01
) {

  trt <- rbinom(n, 1, 0.5)

  # Smoothly changing HR
  HR <- function(t) {
    HR_end - (HR_end - HR_start) * exp(-t / scale)
  }

  # Cumulative treatment hazard
  H_trt <- function(t) {
    lambda * (
      HR_end * t -
      (HR_end - HR_start) * scale *
        (1 - exp(-t / scale))
    )
  }

  event_time <- numeric(n)

  for (i in seq_len(n)) {

    u <- runif(1)
    target <- -log(u)

    if (trt[i] == 0) {

      event_time[i] <- target / lambda

    } else {

      event_time[i] <- uniroot(
        function(t) H_trt(t) - target,
        interval = c(0, 1e4)
      )$root
    }
  }

  censor_time <- rexp(n, rate = censor_rate)

  time <- pmin(event_time, censor_time)
  status <- as.integer(event_time <= censor_time)

  data.frame(time, status, trt)
}
```

## Stacking plots together

```{r}
dat_ph <- sim_ph(n)
fit_ph <- survfit(Surv(time, status) ~ trt, data = dat_ph)

ph_survplot <- ggsurvplot(fit_ph)$plot + 
  labs(title = "Proportional hazards") +
  coord_cartesian(xlim = c(0,60)) + 
  theme_classic(base_size = 15) + 
  theme(legend.position = "none")

dat_cross <- sim_crossing(n)
fit_cross <- survfit(Surv(time, status) ~ trt, data = dat_cross)

crossh_survplot <- ggsurvplot(fit_cross)$plot + 
  labs(title = "Crossed hazards") +
  coord_cartesian(xlim = c(0,60)) + 
  theme_classic(base_size = 15) + 
  theme(legend.position = "none")

dat_delayed <- sim_delayed(n, delay=10, HR_after=0.3)
fit_delayed <- survfit(Surv(time, status) ~ trt, data = dat_delayed)

delay_survplot <- ggsurvplot(fit_delayed)$plot + 
  labs(title = "Delayed or increasing effects") +
  coord_cartesian(xlim = c(0,60)) + 
  theme_classic(base_size = 15) + 
  theme(legend.position = "none")

dat_cure <- sim_cure(n, cure_treatment=0.25)
fit_cure <- survfit(Surv(time, status) ~ trt, data = dat_cure)

ggsurvplot(fit_cure)$plot + 
  labs(title = "Cure effects") +
  coord_cartesian(xlim = c(0,60)) + 
  theme_classic(base_size = 15) + 
  theme(legend.position = "none")

dat_increasing <- sim_increasing_effect(n)
fit_increasing <- survfit(Surv(time, status) ~ trt, data = dat_increasing)

ggsurvplot(fit_increasing)$plot + 
  labs(title = "Increasing hazards") +
  coord_cartesian(xlim = c(0,60)) + 
  theme_classic(base_size = 15) + 
  theme(legend.position = "none")

dat_decreasing <- sim_decreasing_effect(n)
fit_decreasing <- survfit(Surv(time, status) ~ trt, data = dat_decreasing)

dech_survplot <- ggsurvplot(fit_decreasing)$plot + 
  labs(title = "Decreasing hazards") +
  coord_cartesian(xlim = c(0,60)) + 
  theme_classic(base_size = 15) + 
  theme(legend.position = "none")
```

```{r}
ph_vis1 <- ph_survplot + crossh_survplot
ph_vis2 <- ph_survplot + delay_survplot
ph_vis3 <- ph_survplot + dech_survplot
```

```{r}
setwd("C:/Users/vicki/Documents/BTF_repos/stats-survival-analysis")

ggsave("materials/images/ph_vis0.png", plot = ph_survplot,
       width = 4.5, height = 4)

ggsave("materials/images/ph_vis1.png", plot = ph_vis1,
       width = 9, height = 4)

ggsave("materials/images/ph_vis2.png", plot = ph_vis2,
       width = 9, height = 4)

ggsave("materials/images/ph_vis3.png", plot = ph_vis3,
       width = 9, height = 4)
```

```{r}
ph_vis_all <- (ph_survplot + crossh_survplot) / (delay_survplot + dech_survplot)
```

```{r}
ggsave("materials/images/ph_vis_all.png", plot = ph_vis_all,
width = 9, height = 8)
```

## Using smaller sample sizes

Repeat the above, but now with an n of 50

```{r}
set.seed(71)
n <- 50

dat_ph_50 <- sim_ph(n)
fit_ph_50 <- survfit(Surv(time, status) ~ trt, data = dat_ph_50)

ph_survplot_50 <- ggsurvplot(fit_ph_50)$plot + 
  labs(title = "Proportional hazards") +
  coord_cartesian(xlim = c(0,60)) + 
  # change x ticks to match the other plots
  scale_x_continuous(breaks = seq(0, 60, by = 20)) +
  theme_classic(base_size = 15) + 
  theme(legend.position = "none")

ph_survplot + ph_survplot_50
```


```{r}
ph_vis5 <- ph_survplot + ph_survplot_50

setwd("C:/Users/vicki/Documents/BTF_repos/stats-survival-analysis")

ggsave("materials/images/ph_vis5.png", plot = ph_vis5,
       width = 9, height = 4)
```

Figure out why this particular image is coming out smaller than all the others in powerpoint?

```{r}
set.seed(71)
n <- 20

dat_ph_20 <- sim_ph(n)
fit_ph_20 <- survfit(Surv(time, status) ~ trt, data = dat_ph_20)

ph_survplot_20 <- ggsurvplot(fit_ph_20)$plot + 
  labs(title = "Also proportional hazards") +
  coord_cartesian(xlim = c(0,60)) + 
  # change x ticks to match the other plots
  scale_x_continuous(breaks = seq(0, 60, by = 20)) +
  theme_classic(base_size = 15) + 
  theme(legend.position = "none")

ph_survplot + ph_survplot_20
```

```{r}
ph_vis6 <- ph_survplot + ph_survplot_20

setwd("C:/Users/vicki/Documents/BTF_repos/stats-survival-analysis")

ggsave("materials/images/ph_vis6.png", plot = ph_vis6,
       width = 9, height = 4)
```

## Create a "real" survival curve for the proportional hazards example

```{r}
set.seed(71)
n <- 50000

dat_ph_true <- sim_ph(n)
fit_ph_true <- survfit(Surv(time, status) ~ trt, data = dat_ph_true)

ph_survplot_true <- ggsurvplot(fit_ph_true, censor = FALSE)$plot + 
  labs(title = "True population curves") +
  coord_cartesian(xlim = c(0,60)) + 
  # change x ticks to match the other plots
  scale_x_continuous(breaks = seq(0, 60, by = 20)) +
  theme_classic(base_size = 15) + 
  theme(legend.position = "none")

ph_survplot + ph_survplot_true
```


```{r}
ph_vis7 <- ph_survplot + ph_survplot_true

ggsave("materials/images/ph_vis7.png", plot = ph_vis7,
       width = 9, height = 4)
```


```{r}
ph_vis8 <- ph_survplot_true + ph_survplot_20
ggsave("materials/images/ph_vis8.png", plot = ph_vis8,
       width = 9, height = 4)
```


# Visualising baseline hazard rates

Create a series of plots, adding a new hazard rate each time
h(t) on the y axis, t on the x axis

Start with constant hazard

```{r}
constant_hazard <- function(t) {
  lambda <- 0.10
  return(rep(lambda, length(t)))
}

baseline_h <- ggplot(data.frame(t = seq(0, 20, by = 1)), aes(x = t)) +
  stat_function(fun = constant_hazard, color = "#0072CF", linewidth = 1) +
  labs(title = "Constant hazard rate", x = "Time", y = "Hazard rate h(t)") +
  theme_classic(base_size = 15) +
  ylim(0.06, 0.16)

baseline_h
```

Add a decreasing hazard
```{r}
decreasing_hazard <- function(t) {
  lambda <- 0.20
  HR_start <- 0.8
  HR_end <- 0.30
  scale <- 10

  # Smoothly changing HR:
  # HR_start at t = 0
  # approaches HR_end as t increases
  HR <- function(t) {
    HR_end + (HR_start - HR_end) * exp(-t / scale)
  }

  return(lambda * HR(t))
}

baseline_h_2 <- ggplot(data.frame(t = seq(0, 20, by = 1)), aes(x = t)) +
  stat_function(fun = constant_hazard, color = "grey", linewidth = 1) +
  stat_function(fun = decreasing_hazard, color = "#0072CF", linewidth = 1) +
  labs(title = "Decreasing hazard rate", x = "Time", y = "Hazard rate h(t)") +
  theme_classic(base_size = 15)  +
  ylim(0.06, 0.16)

baseline_h_2
```

Increasing hazard
```{r}
increasing_hazard <- function(t) {
  lambda <- 0.20
  HR_start <- 0.30
  HR_end <- 0.8
  scale <- 8

  # Smoothly changing HR
  HR <- function(t) {
    HR_end - (HR_end - HR_start) * exp(-t / scale)
  }

  return(lambda * HR(t))
}

baseline_h_3 <- ggplot(data.frame(t = seq(0, 20, by = 1)), aes(x = t)) +
  stat_function(fun = constant_hazard, color = "grey", linewidth = 1) +
  stat_function(fun = decreasing_hazard, color = "grey", linewidth = 1) +
  stat_function(fun = increasing_hazard, color = "#0072CF", linewidth = 1) +
  labs(title = "Increasing hazard rate", x = "Time", y = "Hazard rate h(t)") +
  theme_classic(base_size = 15) +
  ylim(0.06, 0.16)

baseline_h_3
```

Bathtub shape - increases at first, levels out, then decreases
```{r}
bathtub_hazard <- function(t) {
  hazard_min <- 0.085
  hazard_max <- 0.135
  
  hazard_min + (hazard_max - hazard_min) * ((t - 10) / 10)^6
}

baseline_h_4 <- ggplot(data.frame(t = seq(0, 20, by = 1)), aes(x = t)) +
  stat_function(fun = constant_hazard, color = "grey", linewidth = 1) +
  stat_function(fun = decreasing_hazard, color = "grey", linewidth = 1) +
  stat_function(fun = increasing_hazard, color = "grey", linewidth = 1) +
  stat_function(fun = bathtub_hazard, color = "#0072CF", linewidth = 1) +
  labs(title = "Bathtub-shaped hazard rate", x = "Time", y = "Hazard rate h(t)") +
  theme_classic(base_size = 15) +
  ylim(0.06, 0.16)

baseline_h_4
```

```{r}
ggsave("materials/images/baseline_hazard_rates1.png", plot = baseline_h,
       width = 7, height = 3.5)

ggsave("materials/images/baseline_hazard_rates2.png", plot = baseline_h_2,
       width = 7, height = 3.5)

ggsave("materials/images/baseline_hazard_rates3.png", plot = baseline_h_3,
       width = 7, height = 3.5)

ggsave("materials/images/baseline_hazard_rates4.png", plot = baseline_h_4,
       width = 7, height = 3.5)
```


```{r}
baseline_h_eggs <- ggplot(data.frame(t = seq(0, 20, by = 1)), aes(x = t)) +
  stat_function(fun = increasing_hazard, color = "#0072CF", linewidth = 1) +
  labs(title = "Baseline hazard", x = "Time", y = "Hazard rate h(t)") +
  theme_classic(base_size = 15)  +
  ylim(0.05, 0.5)

baseline_h_eggs
```


```{r}
cox_hazard <- function(t, x, beta) {
  increasing_hazard(t) * exp(beta * x)
}

t <- seq(0, 20, length.out = 200)

beta <- 0.7

plot_data <- data.frame(
  time = rep(t, 4),
  x = factor(rep(c(0, 0.5, 1, 1.5), each = length(t)))
)

plot_data$hazard <- mapply(
  function(time, x) {
    increasing_hazard(time) * exp(beta * as.numeric(as.character(x)))
  },
  plot_data$time,
  plot_data$x
)

baseline_h_eggs_2 <- ggplot(plot_data, aes(x = time, y = hazard, colour = x)) +
  geom_line(linewidth = 1.2) +
  labs(title = "Effect of weight", x = "Time", y = "Hazard rate h(t)", colour = "") +
  theme_classic(base_size = 15) +
  # change colours of lines and labels in legend
  # make the colours different variations of the original blue
  scale_color_manual(values = c("#0072CF", "#49ADFF", "#86C9FF", "#C2E4FF"), 
                     labels = c("Light", "Medium", "Heavy", "Very heavy")) +
  ylim(0.05, 0.5) +
  # move legend to top left corner of plot and make it smaller
  theme(legend.position = c(0.2, 0.9),
        legend.key.size = unit(0.5, "cm"),
        legend.text = element_text(size = 10))

baseline_h_eggs_2
```


```{r}
ggsave("materials/images/baseline_hazard_eggs.png", plot = baseline_h_eggs,
       width = 5, height = 3.5)

ggsave("materials/images/baseline_hazard_eggs2.png", plot = baseline_h_eggs_2,
       width = 5, height = 3.5)
```

# Exercises - log-ranks/survival curves

## Leaves falling

```{r}
set.seed(71)

# Sample size
n <- 43

# Baseline hazard rate (lambda for exponential)
lambda <- 0.05

# Simulate survival times
# T ~ Exponential(rate = lambda)
time <- rexp(n, rate = lambda)

# Random censoring
censoring <- runif(n, min = 0, max = quantile(time, 0.8))
observed_time <- pmin(time, censoring)
event <- as.numeric(time <= censoring)

observed_time <- ceiling(observed_time) # using ceiling rather than round prevents left-censored events

leaves <- data.frame(
  days = observed_time,
  fallen = event
)
```

```{r}
write_csv(leaves, "C:/Users/vicki/Documents/BTF_repos/stats-survival-analysis/materials/data/leaves.csv")
```


## Mechanical parts failures

```{r}
set.seed(71)

# Sample size
n <- 68

# Define true coefficient (log-hazard ratio)
beta_manufacturer <- 0.4     # Manufacturer B has a higher failure hazard

# Covariate
manufacturer <- rbinom(n, 1, 0.5)       # binary (0 = A, 1 = B)

# Linear predictor
lin_pred <- manufacturer * beta_manufacturer

# Baseline hazard rate (lambda for exponential)
lambda <- 0.03

# Simulate survival times
# T ~ Exponential(rate = lambda * exp(lin_pred))
time <- rexp(n, rate = lambda * exp(lin_pred))

# Random censoring
censoring <- runif(n, min = 0, max = quantile(time, 0.8))
observed_time <- pmin(time, censoring)
event <- as.numeric(time <= censoring)

observed_time <- ceiling(observed_time) # using ceiling rather than round prevents left-censored events

parts <- data.frame(
  time = observed_time,
  failed = event,
  manufacturer = factor(manufacturer, labels = c("A", "B"))
)
```

```{r}
write_csv(parts, "C:/Users/vicki/Documents/BTF_repos/stats-survival-analysis/materials/data/parts.csv")
```

## Fruit

Bananas vs oranges
```{r}
set.seed(71)

# Sample size
n <- 38

# Define true coefficient (log-hazard ratio)
beta_fruit <- 0.7     # bananas have a higher hazard of going off

# Covariate
fruit <- rbinom(n, 1, 0.5)       # binary (0 = orange, 1 = banana)

# Linear predictor
lin_pred <- fruit * beta_fruit

# Baseline hazard rate (lambda for exponential)
lambda <- 0.05

# Simulate survival times
# T ~ Exponential(rate = lambda * exp(lin_pred))
time <- rexp(n, rate = lambda * exp(lin_pred))

# Random censoring
censoring <- runif(n, min = 0, max = quantile(time, 0.8))
observed_time <- pmin(time, censoring)
event <- as.numeric(time <= censoring)

observed_time <- ceiling(observed_time)

fruit_bowl <- data.frame(
  time = observed_time,
  gone_off = event,
  fruit = factor(fruit, labels = c("orange", "banana"))
)
```

Bananas, oranges & apples
```{r}
set.seed(71)

# Sample size
n <- 52

# Define true coefficients (log-hazard ratios)
beta_banana <- 0.7       # bananas have the highest hazard
beta_orange <- 0.3       # oranges have an intermediate hazard

# Covariate
fruit <- sample(
  c("apple", "orange", "banana"),
  size = n,
  replace = TRUE
)

# Linear predictor
lin_pred <- ifelse(
  fruit == "banana", beta_banana,
  ifelse(fruit == "orange", beta_orange, 0)
)

# Baseline hazard rate (lambda for exponential)
lambda <- 0.05

# Simulate survival times
# T ~ Exponential(rate = lambda * exp(lin_pred))
time <- rexp(n, rate = lambda * exp(lin_pred))

# Random censoring
censoring <- runif(n, min = 0, max = quantile(time, 0.8))
observed_time <- pmin(time, censoring)
event <- as.numeric(time <= censoring)

observed_time <- ceiling(observed_time)

fruit_bowl_2 <- data.frame(
  time = observed_time,
  gone_off = event,
  fruit = factor(
    fruit,
    levels = c("apple", "orange", "banana")
  )
)
```

```{r}
write_csv(fruit_bowl, "C:/Users/vicki/Documents/BTF_repos/stats-survival-analysis/materials/data/fruit_bowl.csv")
write_csv(fruit_bowl_2, "C:/Users/vicki/Documents/BTF_repos/stats-survival-analysis/materials/data/fruit_bowl_2.csv")
```




