---
title: "00-tests"
editor: visual
---

```{r}
library(survival)
library(survminer)
library(tidyverse)
```

Using lung dataset from survival package - can access all datasets at once using the following command. data(lung) not working for some reason.

```{r}
data(cancer, package="survival")
```

Use survfit function to fit the model. Create survival object using the Surv() function, treat as y, otherwise formula as normal

```{r}
km_lung <- survfit(Surv(time, status) ~ sex, data = lung)

summary(km_lung)
```

What's in the table?!

-   n: total number of subjects used to fit that curve
-   time: the steps/timepoints on the curve
-   n.risk: number of subjects at risk at time t (will start = n)
-   n.event: number of events that occurred at time t
-   n.censor: number of censored subjects (leaving the risk set without event) at time t
-   confidence intervals for the curve
-   strata: stratification of curve estiamtion; if not NULL, there will be multiple curves

```{r}
km_lung_table <- data.frame(time = km_lung$time,
                  n.risk = km_lung$n.risk,
                  n.event = km_lung$n.event,
                  n.censor = km_lung$n.censor,
                  surv = km_lung$surv,
                  upper = km_lung$upper,
                  lower = km_lung$lower
                  )
```

Visualise curve(s):

```{r}
ggsurvplot(km_lung)
```

```{r}
ggsurvplot(km_lung,
           pval = TRUE, conf.interval = TRUE)
```

```{r}
ggsurvplot(km_lung,
           pval = TRUE, conf.interval = TRUE,
           risk.table = TRUE, risk.table.col = "strata")
```

```{r}
ggsurvplot(km_lung,
           pval = TRUE, conf.interval = TRUE,
           risk.table = TRUE, risk.table.col = "strata",
           surv.median.line = "hv")
```

Other additions/changes you could make:

```{r}
ggsurvplot(
   km_lung,                  # survfit object with calculated statistics.
   pval = TRUE,              # show p-value of log-rank test.
   conf.int = TRUE,          # show confidence intervals for 
                             # point estimaes of survival curves.
   conf.int.style = "step",  # customize style of confidence intervals
   xlab = "Time in days",    # customize X axis label.
   break.time.by = 200,      # break X axis in time intervals by 200.
   ggtheme = theme_light(),  # customize plot and risk table with a theme.
   risk.table = "abs_pct",   # absolute number and percentage at risk.
  risk.table.y.text.col = T, # colour risk table text annotations.
  risk.table.y.text = FALSE, # show bars instead of names in text annotations
                             # in legend of risk table.
  ncensor.plot = TRUE,       # plot the number of censored subjects at time t
  surv.median.line = "hv",   # add the median survival pointer.
  legend.labs = 
    c("Male", "Female"),     # change legend labels.
  palette = 
    c("#E7B800", "#2E9FDF")  # custom color palettes.
)
```

Access median survival times:

```{r}
summary(km_lung)$table
```
