---
title: "Evaluating a Cox model"
editor: visual
---

```{r}
#| echo: false
#| message: false
#| results: hide
source(file = "setup_files/setup.R")
```

```{python}
#| echo: false
#| message: false
#| results: hide
exec(open('setup_files/setup.py').read())
import shutup;shutup.please()
```

## Libraries and functions

:::: {.callout-note collapse="true"}
## Click to expand

::: {.panel-tabset group="language"}
## R

```{r, load libraries}
#| eval: false
# load the required packages for fitting & visualising
library(tidyverse)
library(survival)
library(survminer)
```

## Python

```{python, import packages}
#| eval: false
import numpy as np
import pandas as pd
from lifelines import *
import matplotlib.pyplot as plt
from lifelines.statistics import multivariate_logrank_test
from lifelines.statistics import pairwise_logrank_test
```
:::
::::

## Time-varying predictors

It's worth noting that all of the examples we've discussed in this course so far feature predictors that do not vary with time.

For example, whether a patient is male or female, or what age they were when they were admitted to hospital, does not change during the course of the time they're being observed.

In all of our examples, we are also assuming that the *effect* of the predictors on the probability of the event occurring also does not vary with time.

For example, we assume that a patient being male has the same effect on their chance to be discharged from hospital on day 1, as it does on day 30.

These are two related, but distinct ideas:

- Time-varying **predictors**
- Time-varying **effects** of predictors

There will be a bonus chapter later that addresses how to deal with situations where predictors vary with time.

If the effect of a predictor varies with time, this will cause a violation of the proportional hazards assumption, mentioned above. Again, we will talk about this more in a subsequent chapter.

## Summary

Text

::: callout-tip
#### Key Points

- Text
:::
