---
title: "Kaplan-Meier survival curves"
output: html_document
---

```{r, load packages}
#| echo: false
#| message: false
#| results: hide
source(file = "setup_files/setup.R")
```

```{python, import libraries}
#| echo: false
#| message: false
#| results: hide
exec(open('setup_files/setup.py').read())
import shutup;shutup.please()
```

The most common way to present or visualise time-to-event data is using survival curves, fitted with the Kaplan-Meier estimator.

In this chapter, we'll look at how to produce those curves, and some best practices for presenting them in papers or reports.

## Libraries and functions

:::: {.callout-note collapse="true"}
## Click to expand

::: {.panel-tabset group="language"}
## R

```{r, load libraries}
#| eval: false
# load the required packages for fitting & visualising
library(tidyverse)
library(survival)
library(survminer)
```

## Python

```{python, import packages}
#| eval: false
import numpy as np
import pandas as pd

from lifelines import *

import matplotlib.pyplot as plt
```
:::
::::

As our worked example in this chapter, we'll use a dataset called `hospital_stays`.

::: {.panel-tabset group="language"}
## R

```{r, read discharge dataset}
hospital_stays <- read_csv("data/hospital_stays.csv")

head(hospital_stays)
```

## Python

```{python, load discharge dataset}
hospital_stays = pd.read_csv("data/hospital_stays.csv")

hospital_stays.head()
```
:::

In this clinical study, patients were tracked for up to 30 days following hospital admission, to record whether or not they were discharged (and if so, which day following admission).

For each patient, the researcher recorded `age`, `sex`, whether they had `surgery` during their stay, and whether they had a `weekend_admission` (versus a weekday), to assess whether any of these factors have an impact on how quickly they were discharged.

For the purposes of this chapter, we're going to focus just on the `sex` and `surgery` categorical predictors.

## Producing a single survival curve

Let's start by visualising our `hospital_stays` data with a single survival curve.

The first step in this process is to create the survival table, as you saw in the lecture materials.

::: {.panel-tabset group="language"}
## R

In R, we use the `survfit` function to create the survival object.

You will hopefully recognise the syntax - it looks very familiar to how we would write the formula of a linear model, in the `y ~ x` format. (We're not interested in any of our grouping variables just yet, so we're just putting in `1` to look at the whole dataset.)

```{r, create survival object hospitals}
hosp_curve <- survfit(Surv(time, discharged) ~ 1, data = hospital_stays)
```

Incidentally, we could do this in two steps if we wanted - the following code would achieve the same thing:

```{r, multistep create survival object hosp}
#| eval: false

surv_object <- Surv(time, discharged)

hosp_curve <- survfit(surv_object ~ 1, data = hospital_stays)
```

## Python

In the `lifelines` package, we first need to initialise our `KaplanMeierFitter` with some sensible name.

Then, we identify the relevant columns in our dataset that contain the observed times and a record of whether or not the event (in this case, discharge) occurred.

```{python, split discharge dataset for hosp.fit}
hosp = KaplanMeierFitter()

T = hospital_stays["time"]
E = hospital_stays["discharged"]

hosp.fit(T, event_observed=E)
```
:::

Let's look in more detail at the object(s) we just created using our Kaplan-Meier estimation method.

::: {.panel-tabset group="language"}
## R

Our raw data have been reorganised into a new table which has a unique row per time point, instead of a unique row per patient.

We can view this using the `summary` function:

```{r, view hosp survcurve obj}
#| eval: false

summary(hosp_curve)
```

Per time point, we can see the raw numbers of patients who experienced the event, were censored and/or were at risk, plus the Kaplan-Meier-estimated survival probability

## Python

In `lifelines`, the relevant information is stored in two different objects.

The `event_table` gives us the raw numbers at each time point, and the `survival_function_` contains the Kaplan-Meier-estimated survival probability.

```{python, view event/survival table hosp.fit}
hosp.event_table

hosp.survival_function_
```
:::

Now that we've done the Kaplan-Meier estimation step, it's time to plot the survival probabilities against time, to produce the classic survival curve.

::: {.panel-tabset group="language"}
## R

In the `survival` package, there is a purpose-built function for doing this, called `ggsurvplot`.

```{r, simple survcurv hosp}
ggsurvplot(hosp_curve)
```

You can change different aspects of the plot's appearance with the various arguments of the function.

-   The `colour` and `palette` arguments let you control the colour(s)
-   You can edit the axis labels with `ylab` and `xlab`
-   Confidence intervals can (and should!) be explicitly added with the `conf.int` argument

```{r, survcurv hosp optional args}
ggsurvplot(hosp_curve,
          colour="black", 
          palette="#EA7125",
          ylab="Probability of remaining in hospital",
          conf.int=TRUE)
```

## Python

In the `lifelines` package, there's a dedicated function for this called `plot_survival_function`.

```{python, simple curve hosp.fit}
plt.clf()
hosp.plot_survival_function()
plt.show()
```

By default, this will show nothing except the curve with confidence intervals.

However, you can customise your plot by:

-   Adding markers for censored observations (and changing what those markers look like)
-   Changing the colour and line style for the survival curve
-   Editing the x and y axis labels, and adding a title

You can also explicitly set the `ci_show` argument to `True` (though this is the default behaviour).

```{python, curve hosp.fit edited}
plt.clf()

hosp.plot_survival_function(
  ci_show=True,
  color = 'gray',
  linestyle = '--',
  linewidth = 2,
  show_censors=True, 
  censor_styles={"marker": "x", "ms": 8, "mec": "#EA7125"}
)

plt.title("Hospital stays survival curve")
plt.xlabel("Time (days)")
plt.ylabel("Probability of remaining in hospital")

plt.show()
```
:::

And there you are! A survival curve for the `hospital_stays` dataset.

There's still more work to be done, however.

The next few sections will show you how to make the curve more sophisticated, and how to interpret the curve.

### Median survival rate

One of the things to focus on when interpreting a survival curve is the median survival rate (sometimes referred to as the "half-life").

This is the value of `time` on the x-axis when the "survival" probability on the y-axis hits 50%.

It's good practice to include this on your survival curve.

:::: {.panel-tabset group="language"}
## R

In `ggsurvplot`, there is an argument called `surv.median.line` for adding the median.

You can set it to `h`, `v` or `hv` (for vertical, horizontal or both).

```{r, survcurv hosp add median}
ggsurvplot(hosp_curve,
          colour = "black", 
          palette = "#EA7125",
          ylab = "Probability of remaining in hospital",
          conf.int = TRUE,
          surv.median.line = "hv")
```

If we want to figure out what that median value is, we can find it directly in our `hosp_curve` object:

```{r, print median hosp curve}
hosp_curve
```

## Python

There is, unfortunately, no inbuilt function in `lifelines` for adding median survival times to the plot.

You could do this manually, but I have taken the liberty of writing a helper function called `add_median_markers` that adds this functionality. You'll need to run the function code before you can use it.

::: {.callout-note collapse="true"}
#### add_median_markers helper function

```{python, median plotting helper function}
def add_median_markers(fitters, ax=None, short_lines=True, colors=None):
    
    """
    Add median survival markers to a survival plot.

    Parameters:
    - fitters: A single fitter or a list of fitted lifelines KaplanMeierFitter models.
    - ax: Optional matplotlib Axes to draw into.
    - short_lines: If True, draw short lines. If False, draw full-width/height lines.
    - colors: Optional list of colors for the markers (must match number of fitters).
    """
    
    if not isinstance(fitters, list):
        fitters = [fitters]

    if ax is None:
        ax = plt.gca()

    if colors is None:
        # Cycle through default matplotlib colors
        colors = plt.rcParams['axes.prop_cycle'].by_key()['color']

    for i, kmf in enumerate(fitters):
        median = kmf.median_survival_time_
        color = colors[i % len(colors)]
        label = getattr(kmf, 'label', f'Group {i + 1}')

        if short_lines:
            # Short vertical and horizontal lines
            ax.plot([median, median], [0, 0.51], linestyle=':', color=color, label=f'{label} Median: {median:.2f}')
            ax.plot([0, median], [0.5, 0.5], linestyle=':', color=color)
        else:
            # Full lines across the plot
            ax.axvline(median, linestyle='--', color=color, label=f'{label} Median: {median:.2f}')
            ax.axhline(0.5, linestyle=':', color=color)

    ax.legend()
```
:::

```{python, hosp plot median markers}
plt.clf()
hosp.plot_survival_function(show_censors=True)
add_median_markers(hosp, colors = "red")
plt.show()
```

The helper function is set up so that the median value is listed directly in the figure legend, but it's also stored in our `hosp` object and we can access it like this:

```{python, print hosp median}
print(hosp.median_survival_time_)
```
::::

We can see that the median survival time for our `hospital_stays` dataset is 9.5.

::: callout-tip
#### Interpreting median survival time

The median survival time represents the time point at which we can expect 50% of individuals to have experienced the event (and 50% not to have experienced it).

It is essentially a descriptive statistic for our survival curve.

When we have more than one curve (later in this chapter),/pre we can use the median survival as a point of comparison. With just one curve, it still provides us with a general expectation of how long patients can expect to spend in hospital.

We use the median, rather than the mean, because the median is less affected by skewed data and outliers (which are both common in time-to-event data).

In this dataset, that means that we expect (on average) that 50% of people will be discharged within the first 9.5 days of being admitted to hospital.
:::

### Adding risk table

The median survival time by itself isn't always super informative, because a) by definition it only refers to one point in the curve, and b) it doesn't give any indication of sample size.

We can make our survival curve more informative by including risk or cumulative event tables underneath.

This is especially important if you plan to present or share your survival curve.

::: {.panel-tabset group="language"}
## R

```{r, hosp curve with risk table}
ggsurvplot(hosp_curve,
          colour = "black", 
          palette = "#EA7125",
          ylab = "Probability of remaining in hospital",
          conf.int = TRUE,
          surv.median.line = "hv",
          risk.table = TRUE)
```

## Python

```{python, hosp.curve with risk table}
from lifelines.plotting import add_at_risk_counts

plt.clf()

fig, ax = plt.subplots()

hosp.plot_survival_function(ax=ax, show_censors=True)
add_median_markers(hosp, ax=ax, colors = "red")
add_at_risk_counts(hosp, ax=ax, fig=fig)

plt.subplots_adjust(bottom=0.3)
plt.show()
```
:::

### Adding cumulative events table

This is an alternative (or addition) to the risk table, albeit less common.

::: {.panel-tabset group="language"}
## R

In `ggsurvplot`, you can show either or both:

```{r, hosp curve with cum ev table}
ggsurvplot(hosp_curve,
          colour = "black", 
          palette = "#EA7125",
          ylab = "Probability of remaining in hospital",
          conf.int = TRUE,
          surv.median.line = "hv",
          risk.table = TRUE,
          cumevents = TRUE)
```

## Python

The `lifelines` package unfortunately doesn't have a function for adding a cumulative events table, and it's a bit of a pain to do it manually. (Though if you're handy with `matplotlib` etc., you're most welcome to try!)

As an intermediate measure, you can at least extract the cumulative events table to look at/present separately:

```{python, showing cum ev counts}
cumulative_events = hosp.event_table['observed'].cumsum()

print(cumulative_events)
```
:::

## Stratified survival curves

Now that we know how to plot a single survival curve, let's explore stratified curves - where we split the data up according to categorical/grouping variables.

We'll focus on two of the categorical variables from `hospital_stays`: `surgery` and `sex`.

### Stratify by surgery

"Stratify" just means split according to category.

Our `surgery` variable is binary: each patient either received surgery, or did not. We want to know whether those two subgroups have different curves.

::: {.panel-tabset group="language"}
## R

We barely have to change the code in R to achieve this. We just have to specify that the survival is being predicted by `sex` in our `survfit` formula:

```{r, survfit surgery strat}
hosp_curve_surgery <- survfit(Surv(time, discharged) ~ surgery, 
                              data = hospital_stays)

ggsurvplot(hosp_curve_surgery,
           conf.int = TRUE,
           surv.median.line = "hv")
```

The stratification will automatically be extended to our risk or cumulative events tables, too, if we add those to our plot.

```{r, plot surgery strat}
ggsurvplot(hosp_curve_surgery,
           conf.int = TRUE,
           surv.median.line = "hv",
           risk.table = TRUE,
           cumevents = TRUE)
```

## Python

This first portion is a copy of the code we did above, when we created a single non-stratified curve. You may not need to run this again.

```{python, hosp T/E}
T = hospital_stays["time"]
E = hospital_stays["discharged"]
```

Now, we need to create two separate Kaplan Meier fitted objects for our two different groups. To achieve this, we need to subset our data by the `surgery` variable.

```{python, }
# Subset of the dataset that received surgery
surg = (hospital_stays["surgery"] == 1)

# Kaplan Meier fit for surgical data subset
surg_hosp = KaplanMeierFitter()
surg_hosp.fit(T[surg], event_observed=E[surg], label="Surgical patients")

# Kaplan Meier fit for non-surgical data subset
ns_hosp = KaplanMeierFitter()
ns_hosp.fit(T[~surg], event_observed=E[~surg], label="Non-surgical patients")
```

Now, we can plot both of these curves on the same set of axes.

Our `add_median_markers` helper function can be used again here - we just have to make sure to pass both of the fit objects, as a list, as the first argument.

```{python}
plt.clf()

# Initialise a plot to capture both of the curves
fig, ax = plt.subplots()

# Add the two curves, one at a time
surg_hosp.plot_survival_function(ax = ax)
ns_hosp.plot_survival_function(ax = ax)

# Add the median markers for both curves
add_median_markers([surg_hosp, ns_hosp], ax = ax)

plt.show()
```
:::

So, what do we see when we compare surgical and non-surgical patients?

The two curves look very different.

The non-surgical group has a curve that drops more quickly towards the x-axis. This indicates that more events - here, that means discharges - are happening earlier for the non-surgical group.

This is borne out by our median survival values. Our non-surgical patients have a lower median survival value, suggesting that on average we expect surgical patients to stay longer in hospital before discharge.

Overall, this gives a strong impression that surgical patients stay longer in hospital, i.e., have a longer time to discharge.

Next chapter, we'll look at quantifying this with a significance test.

### Stratify by sex

Another categorical variable in our dataset is biological `sex`. Patients in this dataset were grouped as either male (1) or female (0).

::: {.panel-tabset group="language"}
## R

Once again, there's very little tweaking to the code required to achieve this:

```{r}
hosp_curve_sex <- survfit(Surv(time, discharged) ~ sex, data = hospital_stays)

ggsurvplot(hosp_curve_sex,
           conf.int = TRUE,
           surv.median.line = "hv")
```

```{r}
hosp_curve_sex
```

## Python

We use the same procedure as above, creating a separate fit for the male and female subsets of the data.

```{python}
# Subset the data
female = (hospital_stays["sex"] == 0)

# Kaplan Meier fit for female patients
fem_hosp = KaplanMeierFitter()
fem_hosp.fit(T[female], event_observed=E[female], label="Female patients")

# Kaplan Meier fit for male patients
male_hosp = KaplanMeierFitter()
male_hosp.fit(T[~female], event_observed=E[~female], label="Male patients")
```

Then, we plot both of these curves on the same set of axes, with median markers if desired:

```{python}
plt.clf()

# Initialise a plot
fig, ax = plt.subplots()

# Add both survival curves
fem_hosp.plot_survival_function(ax = ax)
male_hosp.plot_survival_function(ax = ax)

# Add both median markers
add_median_markers([fem_hosp, male_hosp], ax = ax)

plt.show()
```
:::

When stratified by `sex`, the curves look much more similar to each other compared to when we stratified by `surgery`.

The median survival times are also much more similar (9.5 versus 8.5). Female patients get discharged a little quicker on average, but not very much, and the confidence intervals for the two curves are overlapping pretty much the whole way.

Visually, this suggests that `sex` has much less of an impact than `surgery`.

Next chapter we'll look at how to test this more formally.

### Stratify by sex and surgery

::: {.panel-tabset group="language"}
## R

We could stratify our plot in one of two different ways.

Firstly, we could consider our data as containing four separate subgroups.

We use the `strata` syntax inside our `survfit` function, and we will get four separate survival curves in our figure.

```{r}
hosp_curve_sexsurg <- survfit(Surv(time, discharged) ~ strata(sex, surgery), data = hospital_stays)

ggsurvplot(hosp_curve_sexsurg,
           conf.int = TRUE,
           surv.median.line = "hv")
```

Alternatively, we could facet our plot. If we're doing this, we explicitly *don't* use the `strata` syntax, because we want each variable to be estimated separately.

We also need to add the `facet.by` argument into our `ggsurvplot` function:

```{r}
hosp_curve_sexsurg <- survfit(Surv(time, discharged) ~ sex + surgery, data = hospital_stays)

ggsurvplot(hosp_curve_sexsurg,
           conf.int = TRUE,
           facet.by = "sex",
           surv.median.line = "hv")
```

## Python

We can produce four separate survival curves in our figure.

First, we need to create a new grouping variable (this acts a bit like an interaction term):

```{python}
# Combine two binary variables into a single stratification group
hospital_stays["group"] = hospital_stays["sex"].astype(str) + "_" + hospital_stays["surgery"].astype(str)

# Optional: label the groups nicely
hospital_stays["group"] = hospital_stays["group"].map({
    "0_0": "female_nonsurgical",
    "0_1": "female_surgical",
    "1_0": "male_nonsurgical",
    "1_1": "male_surgical"}
)

hospital_stays.head()
```

Then, we need to fit separate Kaplan-Meier estimates for each of our four subgroups.

For efficiency of code, I'm going to use a `for` loop to achieve this (without saving the fitted objects anywhere).

```{python}
hosp = KaplanMeierFitter()

# Initialise a plot
fig, ax = plt.subplots()

# Loop through each subgroup, fitting & plotting the curve
for name, grouped_df in hospital_stays.groupby("group"):
    hosp.fit(grouped_df["time"], event_observed=grouped_df["discharged"], label=name)
    hosp.plot_survival_function(ax=ax)
    add_median_markers(hosp, ax=ax, colors="k")
    
plt.show()
```

It is possible to achieve a faceted plot with some clever additions from `seaborn`, but the code gets quite extensive, so we'll skip that for now.
:::

## Best practices for figures

You might be using Kaplan-Meier curves like these simply to visualise, explore and understand your own data. However, it's very likely that you're hoping to use curves like these to present time-to-event data in scientific writing.

If so, here are some best practices to follow, to be clear and transparent with your figure(s).

+---+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| ✓ | **Always include confidence intervals**                                                                                                                                                             |
+---+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|   | At best, a survival curve without confidence intervals is just a bit less informative; at worst, it's actively disingenuous.                                                                        |
|   |                                                                                                                                                                                                     |
|   | There's no reason *not* to include the intervals - they are only ever helpful/informative!                                                                                                          |
+---+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| ✓ | **Include & report the median survival time**                                                                                                                                                       |
+---+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|   | It's a small addition to the plot (you can make it visually quite unobtrusive, by making the median lines a bit fainter), but adds quite a bit of helpful info to a viewer.                         |
+---+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| ✓ | **Include summary table with counts**                                                                                                                                                               |
+---+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|   | This is more commonly a risk table, but you might find the cumulative events table (or both) more useful. In either case, they give context to the curve by making the sample size(s) more obvious. |
+---+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| ✓ | **Make sure the axes labels are clear**                                                                                                                                                             |
+---+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|   | Consider changing the y-axis label from "survival probability" to something more appropriate for your particular research question.                                                                 |
|   |                                                                                                                                                                                                     |
|   | If death is not the event of interest in your dataset, then "survival" is probably not the most helpful label.                                                                                      |
+---+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| ✓ | **Show relevant p-values**                                                                                                                                                                          |
+---+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|   | It's common and often helpful to add the p-value to your survival curves, for readers' ease.                                                                                                        |
|   |                                                                                                                                                                                                     |
|   | In subsequent chapters, we'll look at how to calculate the p-values, and show how they can be added to the plots.                                                                                   |
+---+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

## Summary

One of the most common ways to visualise time-to-event (survival) data is via Kaplan-Meier curves. These are easily fitted using dedicated packages in R/Python.

::: callout-tip
#### Key Points

-   A Kaplan-Meier survival curve plots `time` against the estimated `survival probability` (calculated using the Kaplan-Meier estimator)
-   You can fit a Kaplan-Meier curve for the entire dataset, or stratify by one or more categorical predictors
-   You should always include confidence intervals for your curves
-   Other helpful features include the median survival time and risk tables/cumulative event tables
:::
