---
title: "Two or more predictors"
editor: visual
---

```{r}
#| echo: false
#| message: false
#| results: hide
source(file = "setup_files/setup.R")
```

```{python}
#| echo: false
#| message: false
#| results: hide
exec(open('setup_files/setup.py').read())
import shutup;shutup.please()
```

::: callout-warning
This chapter is still in development.
:::

In the previous sections, we focused on log-rank tests for single categorical predictors, and Cox regression for single continuous predictors.

However, just like in a regular linear model, a Cox regression can include multiple predictors. We'll explore what that looks like in this chapter.

## Libraries and functions

:::: {.callout-note collapse="true"}
## Click to expand

::: {.panel-tabset group="language"}
## R

```{r, load libraries}
#| eval: false
# load the required packages for fitting & visualising
library(tidyverse)
library(survival)
library(survminer)
```

## Python

```{python, import packages}
#| eval: false
import numpy as np
import pandas as pd
from lifelines import *
import matplotlib.pyplot as plt
from lifelines.statistics import multivariate_logrank_test
from lifelines.statistics import pairwise_logrank_test
```
:::
::::

## Fitting multivariate models

## Visualising models

## Checking assumptions

## Interpreting the model

Much the same as with a single continuous predictor

You have:

- hazard ratios

- global significance tests

- individual significance tests

## Making predictions

## Summary

Text

::: callout-tip
#### Key Points

- Text
:::
