---
title: "Two or more predictors"
editor: visual
---

```{r}
#| echo: false
#| message: false
#| results: hide
source(file = "setup_files/setup.R")
```

```{python}
#| echo: false
#| message: false
#| results: hide
exec(open('setup_files/setup.py').read())
import shutup;shutup.please()
```

In the previous sections, we focused on log-rank tests for single categorical predictors, and Cox regression for single continuous predictors.

However, just like in a regular linear model, a Cox regression can include multiple predictors. We'll explore what that looks like in this chapter.

::: callout-warning
**This chapter is still in development.**

Note that this chapter only features code in R. We hope to update it in future to include Python.
:::

## Libraries and functions

:::: {.callout-note collapse="true"}
## Click to expand

::: {.panel-tabset group="language"}
## R

```{r, load libraries}
#| eval: false
# Load the required packages for fitting & visualising
library(tidyverse)
library(survival)
library(survminer)
```

## Python

```{python, import packages}
#| eval: false
import numpy as np
import pandas as pd
from lifelines import *
import matplotlib.pyplot as plt
from lifelines.statistics import multivariate_logrank_test
from lifelines.statistics import pairwise_logrank_test
```
:::
::::

## Fitting models with 2+ predictors

The good news about Cox regression is, just like regular linear models, once you know the syntax for one predictor, it's very quick and easy to add others!

Using our trusty `hospital_stays` dataset, let's recreate our one-predictor Cox regression (with `age` as the single continuous predictor).

Then, let's follow it up with a three-predictor model, where we include `surgery` and `weekend_admission` simultaneously alongside `age`.

(These two additional variables capture whether or not a patient received surgery, and whether they were admitted on a weekday versus a weekend.)

::: {.panel-tabset group="language"}
## R

The syntax for adding additional predictors is exactly the same as with the `lm` function - we use a plus sign.

```{r, fitting two predictors}
#| message: false

hospital_stays <- read_csv("data/hospital_stays.csv")

# Single continuous predictor
cox_hosp_age <- coxph(Surv(time, discharged) ~ age,
                      data = hospital_stays)

# Two predictors
cox_hosp_multi <- coxph(Surv(time, discharged) ~ age + surgery + weekend_admission,
                        data = hospital_stays)
```
:::

What differences do we observe when we look at the two models?

::: {.panel-tabset group="language"}
## R

```{r, summarising one vs three predictors}
summary(cox_hosp_age)

summary(cox_hosp_multi)

```
:::

The most obvious difference is that, in our multi-predictor model, we get information about `age`, `surgery` and `weekend_admission` in the output.

We also see that some of the numbers have changed slightly. For instance, the concordance value increases a bit when we add more predictors.

(This is actually expected - in the same way that $R^2$ increases when you add more predictors in regular linear models.)

::: callout-tip
### Good news!

It doesn't matter what order you put the predictors into the model. Due to the way that the model is estimated, you'll always get the same coefficients no matter what.
:::

### Interactions

Along with adding more main effects, we can also include interaction terms in our Cox regression model.

::: {.panel-tabset group="language"}
## R

Once again, the syntax is exactly the same as with regular linear models.

```{r, interaction term hosp stays}
cox_hosp_int <- coxph(Surv(time, discharged) ~ age * surgery + weekend_admission,
                        data = hospital_stays)

summary(cox_hosp_int)
```
:::

Here, we have added just one interaction, on top of our existing main effects: `age:surgery`.

We see that the additional term now appears in our output, and that several of the other numbers have changed slightly as well (concordance, p-values, coefficients etc.)

Okay - so now you know how to fit these models. But that's the easy part.

How can we now **evaluate and interpret** the new models we've created?

## Checking assumptions

The first step, as with any type of modelling, is to check whether the assumptions are met.

Our four assumptions for Cox regression from previous chapters still apply:

- Independence
- Non-informative censoring
- Linearity (for continuous predictors)
- Proportional hazards

We actually already considered the [linearity](08-evaluating-cox-model.qmd#sec-linearity-hospstays-1) assumption in a previous chapter for the `age` predictor. Since we haven't added any other continuous predictors, we don't need to construct any more Martingale residuals plots here.

We also already examined the [proportional hazards](08-evaluating-cox-model.qmd#sec-prophaz-hospstays-1) assumption in a previous chapter for the `age` and `surgery` variables.

So, from here, all we really need to do is check is the proportional hazards assumption for the `weekend_admission` variable, and consider any possible sources of non-independence or informative censoring.

#### Proportional hazards: `weekend_admission`

Now, the Schoenfeld residuals for proportional hazards:

::: {.panel-tabset group="language"}
## R

```{r, schoenfeld check for cox_hosp_int}
ggcoxdiagnostics(cox_hosp_int, type = "schoenfeld")
```
:::

No obvious issues are jumping out!

Remember, for our categorical predictors, it's usually also helpful (and easier) to examine the log-log curves. Here's a reminder of how to do that for the `weekend_admission` variable:

::: {.panel-tabset group="language"}
## R

```{r, loglog for wkndadmission}
surv_hosp_adm <- survfit(Surv(time, discharged) ~ weekend_admission,
                             data = hospital_stays)

plot(surv_hosp_adm, fun = "cloglog")
```
:::

This looks pretty parallel to me. (You could also double-check what you think of the actual survival curves, too, if you want some practice.)

#### Possible sources of non-independence/informative censoring

Particularly with the introduction of the `surgery` and `weekend_admission` variables, we can consider a few different possible issues.

| Problem | Assumption affected |
|-----------------------------------------------|-------------------------|
| Patients operated on by the same surgeon/surgical teams may have more similar outcomes | Independence |
| Patients admitted around the same time may have more similar ailments and therefore more similar outcomes (e.g., drinking-related injuries are more common on weekends) | Independence |
| Patients requiring complex or specialist surgeries may be more likely to be transferred to other hospitals, and thus lost from the study | Non-informative censoring |
| Patients admitted at the weekend are more likely to be transferred to another hospital due to low/insufficient staffing, and thus lost from the study | Non-informative censoring |

Perhaps none of these affect our dataset, and from here, we'll assume that they don't. These are just some examples to continue to get you used to thinking about these assumptions.

## Exercises

### Website purchases {#sec-exr-web-purchases}

::::::: callout-note
{{< level 2 >}}

Read in the `website_purchases.csv` file.

These data were collected by a marketing team who are trying to increase sales via their company website. Each row of the .csv file represents a unique customer visiting the website, and the time until they make a purchase was tracked, up to a maximum of 40 minutes.

The marketing team know what type of `device` is being used (mobile or laptop) and want to know whether there is a difference between the two.

For some website visitors, they offer a `discount`, and they want to know whether it has an impact on purchases, and whether this impact differs between mobile and laptop users.

For this exercise, answer the following questions:

1.  What Cox regression model would best address the research question? (Also: fit it!)
2.  Are the assumptions of the model well-met?

::: {.callout-tip collapse="true"}
##### When thinking about assumptions...

Note that right-censoring in this experimental design could occur because a visitor goes over the 40 minute mark, or because they leave the website without making a purchase within 40 minutes.

Think about why might they leave before the 40 minute mark without purchasing. The reason matters when considering certain assumptions!
:::

::::: {.callout-answer collapse="true"}
#### Worked answer

##### Fit a Cox model

The trick here is that the research question asks whether the effect of `discount` differs between the two `device` types – meaning we are interested in a possible interaction.

Therefore, the only correct model that would address this question is a full model with the interaction term included:

::: {.panel-tabset group="language"}
## R

```{r, fit websitepurchases cox reg}
#| message: false
#| warning: false
website_purchases <- read_csv("data/website_purchases.csv")

cox_purchases <- coxph(Surv(time, purchased) ~ device * discount,
                       data = website_purchases)

summary(cox_purchases)
```
:::

We'll talk more next chapter about how to interpret this in a little more detail, and return to this model in one of the associated exercises at the end of the chapter.

##### Check assumptions

###### Linearity

There are no continuous predictors in the model, so this assumption doesn't apply.

###### Proportional hazards

To assess this, given that we have two categorical predictors, let's use the log-log survival curves.

We'll need to produce the curves for each predictor individually:

::: {.panel-tabset group="language"}
## R

```{r, check prop haz website purchases}
# Refit using survfit
surv_purchases_device <- survfit(Surv(time, purchased) ~ device,
                       data = website_purchases)

# Produce log-log curves
plot(surv_purchases_device, fun = "cloglog")

# Repeat for discount predictor
surv_purchases_discount <- survfit(Surv(time, purchased) ~ discount,
                                 data = website_purchases)

plot(surv_purchases_discount, fun = "cloglog")
```
:::

They both look alright!

###### Independence

As far as the dataset is described to us, there are no obvious issues here.

However, possible sources of non-independence that could creep in (and which you'd probably want to check for, if you were actually running this analysis in real life) could include:

- What if the same person visits the website multiple times? (For instance, if they decide to switch between devices!)
- What if visitors arrive in groups (e.g., friends/family members browsing together), particularly if they are only planning to make one purchase between them?

###### Non-informative censoring

This, in my opinion, is the most likely assumption to be violated in this design, and it's all because of one of the causes of right-censoring: a visitor leaving the website before they make a purchase.

Why might they leave the website without a purchase?

- They close the browser window due to disinterest
- Their internet connection drops
- They decide to switch devices – note that this would be an independence issue, described just above

And of course you may think of others; this list is not exhaustive.

We *absolutely* care about the reason that the censoring occurred.

If the censoring is due to disinterest, this could be a problem. People who are less interested in purchasing are probably more likely to close the browser window. In other words, the "risk" of experiencing the event (in this case, making a purchase) is related to the likelihood of being censored, and our non-informative censoring assumption may be violated.

On the other hand, censoring due to technical issues is much more likely to be non-informative – technical issues are more likely to be totally random.
:::::
:::::::

### Penguin chicks {#sec-exr-penguin-chicks}

:::: callout-exercise
{{< level 2 >}}

In this dataset, scientists running a conservation programme for blue penguins (also known as the little penguin, or the fairy penguin) track how long it takes for eggs to hatch.

For each egg, they record:

- The `mass` of the egg
- `parent_experience`, i.e., whether the parents are first-time or repeat parents
- The average ambient `temperature` during the incubation

You should:

1.  Fit at least one Cox regression model to the dataset
2.  Check the model assumptions

::: {.callout-answer collapse="true"}
#### Worked answer

Worked answer coming soon.
:::
::::

## Summary

As with a regular linear model, a Cox regression model can have as many predictors as you'd like, of any kind (including interaction terms).

While adding additional terms to the model doesn't make it much harder to fit the model itself, nor does it affect which assumptions apply to the model/data, it can make for a slightly longer assumption-checking procedure.

::: callout-tip
#### Key Points

- Fitting Cox regression models with 2+ predictors (including interactions) follows the same simple syntax as fitting linear models with multiple predictors
- Assumptions of proportional hazards and linearity are checked for each predictor individually/one by one, even when there are multiple predictors
:::
